---
title: "Work with agents"
description: >
  Give an agent bounded reads, a pinned accepted boundary, and a proposal path
  that still passes through validation, provenance, and review.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work with agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

An agent connected to a database can read anything the connection allows, write
without review, and give a different answer tomorrow because the underlying
rows moved. Graft is built so that none of those three things has to be true.

The same properties that make accepted knowledge reviewable for people make it
usable by a model: reads are bounded and deterministic, an accepted boundary
can be pinned and carried between sessions, and every write is a proposal that
carries its producer and passes validation before it is accepted.

| What the agent does | Graft surface | Boundary |
|---|---|---|
| Reads records, searches, traverses, reads history | `graft_tools()` | Read-only, bounded, no SQL or filesystem |
| Works from fixed accepted state | `graft_snapshot()`, `graft_at()` | Later commits cannot change its answers |
| Proposes new or corrected records | `graft_plan()`, `graft_commit()` | Validated, attributed, reviewable before acceptance |
| Edits readable files | `graft_status()`, `graft_review()` | File edits stay proposals until reviewed |

## Set up a store to work against

This guide uses the resolved data-dict contract included with the package.

```{r setup}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "team-directory.data-dict.json",
  package = "graft",
  mustWork = TRUE
))
store <- graft_open(schema, ":memory:", okf = "disabled")

graft_ingest(
  store,
  list(
    organization = data.frame(
      id = "org:daily-planet",
      name = "Daily Planet"
    ),
    person = data.frame(
      id = "person:lois-lane",
      full_name = "Lois Lane",
      job_title = "Reporter"
    ),
    employment = data.frame(
      id = "employment:lois-lane:daily-planet",
      person_id = "person:lois-lane",
      organization_id = "org:daily-planet"
    )
  ),
  graft_provenance(
    producer = "team-directory-import",
    idempotency_key = "team-directory-v1"
  )
)
```

## Hand the agent bounded tools

Install ellmer 0.5.0 or later for `graft_tools()` and `graft_verify()`. The
[compatibility guide](compatibility.html) records the tested versions and the
unpinned Commons setup.

`graft_tools()` returns [ellmer](https://ellmer.tidyverse.org/) tool
definitions for `graft_find()`, `graft_get()`, `graft_query()`, and
`graft_history()`, plus `graft_dictionary()` for data-dict contracts:

```{r tools}
tools <- graft_tools(store)
names(tools)
#> [1] "graft_find"       "graft_get"        "graft_query"
#> [4] "graft_history"    "graft_dictionary"
```

Register them with a chat and the model can search, retrieve, traverse, and
read history on its own:

```{r chat}
chat <- ellmer::chat_anthropic()
chat$set_tools(tools)

chat$chat("Who works at the Daily Planet, and has that person's title changed?")
```

The definitions delegate to the same public retrieval functions a person calls
from the console. They accept no SQL, no connection, no path, and no mutation
argument, and each one is annotated read-only, non-destructive, idempotent, and
closed-world for hosts that act on those hints.

When the accepted ledger contains `GraftDefinition` records, `graft_tools()`
also supplies `graft_definitions` and one `graft_calculate` tool. The first
discovers metrics, filters, derived values, dependencies, and eligible columns.
The second combines same-table definitions and public dimensions at one exact
boundary; it never accepts SQL.

Every tool result is a named list with `result`, `truncated`, `limit`, and one
canonical `receipt`. A model that receives a bounded prefix is told that it
received one, rather than silently reasoning over a partial answer:

```{r result-shape}
str(tools$graft_find(query = "Lois", class = "person", limit = 5), max.level = 2)
#> List of 4
#>  $ result   :'data.frame': 1 obs. of 5 variables:
#>  $ truncated: logi FALSE
#>  $ limit    : int 5
#>  $ receipt  :List of 3
#>   ..$ store   :List of 1
#>   ..$ boundary:List of 4
#>   ..$ schema  :List of 2
```

The receipt names the store, accepted batch and commit order, and both schema
digests. Live tools pin that state for one invocation; tools built from a
`GraftView` also carry its immutable snapshot identifier. A calculation result
adds the complete accepted definition closure under `receipt$definitions`.

## Discover meaning before reading records

Data-dict-backed stores expose `graft_dictionary()` to R callers and as an
ellmer tool. LinkML stores retain their existing tools and report that a
data-dict contract is required if dictionary discovery is called directly.

These calls need no model or API key:

```{r dictionary-discovery}
view <- graft_at(store, graft_snapshot(store))
context <- graft_dictionary(view, table = "person", field = "full_name")
context$result$entries

tools <- graft_tools(view)
metadata <- tools$graft_dictionary(table = "person", field = "full_name")
people <- tools$graft_find(query = "Lois", class = "person", limit = 5)
identical(metadata$receipt, people$receipt)
#> [1] TRUE
```

The entries distinguish supported contract properties such as requiredness
from descriptive labels, units, relationships, glossary terms, and assertions.
Unsupported profile semantics are identified explicitly. Assertion prose is
never executed. Accepted definitions remain available through
`graft_definitions()` rather than being copied from dictionary source text.

Use `limit` and the returned `result$next_offset` for subsequent pages. Each
page contains at most 100 entries, and each string cell is capped at 2,000
characters. `text_truncated` identifies clipped entries; the outer `truncated`
flag reports clipping or another page. A selected field still includes dataset,
table, glossary, and adapter semantics needed to interpret it.

Restricted columns and relationships with restricted endpoints are omitted.
Examples, observed ranges, source locators, and arbitrary extension metadata
are excluded. Public prose is not content-scrubbed: authors must keep private
information out of public descriptions, assertions, and glossary entries.
Relationships report resolved public endpoint pairs and cardinality. Pairs do
not encode operators or aliases, so discovery does not reconstruct joins.
Assertions likewise require resolved public column references;
dataset assertions and assertions without resolved references are omitted.

Dictionary discovery follows the generic-read verification rule. Quoting its
returned prose can support a `cited` answer; an uncited discovery call fails
closed. Combining it with a calculation cannot produce `verified` evidence.
A pinned view retains the same dictionary and boundary receipt after later
commits.

## Use the same boundary with Commons

`graft_commons_data_source()` copies selected public class tables, applicable
normalized relations, and accepted definitions into a detached source created
through Commons's public API:

```{r commons, eval = FALSE}
commons_source <- graft_commons_data_source(view, classes = "person")
layer <- commons::semantic_layer(commons_source)
```

Commons owns that copy, its file measures, and its fallback query paths. It
never receives Graft's live DuckDB connection, and later Graft commits do not
change an existing Commons source.

## Verify recorded answers offline

After the host has recorded a chat, `graft_verify()` classifies its completed,
text-bearing assistant answers from the trace already held by ellmer. It is a
deterministic, read-only inspection: it does not call the model, use the
network, reopen the store, or authenticate identifiers in a receipt.

```{r verify}
verification <- graft_verify(chat)
verification[, c("answer_index", "turn_index", "answer_text", "label")]
verification$reason_codes
verification$diagnostics
```

The result has one row per answer, excluding tool-only turns and partial
answers. Each row includes the answer text, its label, stable reason codes, the
receipts and paired tool calls considered, matched citations, and any
diagnostics.

The three labels classify the recorded evidence path:

- **Verified** (`"verified"`) means the window contains only successful
  `graft_calculate` calls with valid governed receipts.
- **Cited** (`"cited"`) means every successful generic Graft result is
  independently matched to an explicit quotation or Markdown blockquote in
  the answer. A generic read caps mixed calculation and generic evidence at
  this label.
- **Untrusted** (`"untrusted"`) covers missing evidence, unmatched generic
  reads, non-Graft tools, tool errors, malformed receipts, and unsupported
  trace shapes.

Citation candidates are normalized for whitespace, Markdown emphasis,
typographic quotation marks, and dashes. They must contain at least ten
characters and match a textual value in the bounded result with a
case-sensitive fixed match. Unquoted overlap, receipt fields, tool arguments,
and unrelated chat text are not citation evidence.

This is deterministic provenance classification, not fact-checking or
cryptographic authentication. `graft_verify()` does not prove that every claim
follows from the cited text, and it does not reopen a store to authenticate
receipt identifiers.

Diagnostics are separate from trust. For example, valid calculation results
from mixed accepted boundaries are reported in `diagnostics`, but do not
downgrade an otherwise valid `"verified"` label.

## Pin the boundary the agent reasons over

A long-running agent session should not change its mind because another
process committed a batch halfway through. `graft_snapshot()` captures the
accepted boundary as a serializable, path-free value, `graft_at()` binds it to
a read-only view, and tools built from that view stay pinned:

```{r pin}
snapshot <- graft_snapshot(store)
view <- graft_at(store, snapshot)
pinned_tools <- graft_tools(view)
```

Accept a change after pinning:

```{r later-commit}
graft_ingest(
  store,
  list(person = data.frame(
    id = "person:lois-lane",
    full_name = "Lois Lane",
    job_title = "Investigative reporter"
  )),
  graft_provenance(
    producer = "hr-review",
    idempotency_key = "hr-review-v1"
  )
)

graft_get(store, "person:lois-lane")$record$job_title
#> [1] "Investigative reporter"
graft_get(view, "person:lois-lane")$record$job_title
#> [1] "Reporter"
```

Because the snapshot carries no connection and no filesystem path, it can be
written to a job record, passed to another process, or stored with an
evaluation case and rebound later with `graft_at()`. Two runs against the same
snapshot read the same accepted knowledge. `graft_view_snapshot()` recovers the
exact boundary a view is holding, which is what you record alongside a
transcript when you need to explain an answer.

The live-store `integrity` operation is unavailable through a view, since a
diagnostic of the current store would contradict the pinned boundary.

## Let the agent propose, not write

Nothing about agent-supplied records is special: they are ordinary data frames
that go through the same plan, review, and commit path as any other producer.
What matters is that the agent is named as the producer and that the plan is
inspected before it is accepted.

```{r propose}
proposal <- graft_provenance(
  producer = "directory-agent",
  version = "2026.08",
  run_id = "run-8842",
  idempotency_key = "run-8842-batch-1",
  metadata = list(model = "claude-opus-4-5", session = "s-1174")
)

plan <- graft_plan(
  store,
  list(person = data.frame(
    id = "person:clark-kent",
    full_name = "Clark Kent",
    job_title = "Reporter"
  )),
  proposal
)

plan@valid
plan@changes[, c("class", "record_id", "action", "changed_fields")]
plan@issues[, c("class", "record_id", "field", "message")]
```

Planning writes nothing. The plan is an immutable object bound to the store,
schema, and record heads observed during planning, so a person or a host policy
can read `@changes` and `@issues`, decide, and then call `graft_commit()`.
If the store moved underneath the plan, the commit fails its preconditions
instead of writing a decision that was reviewed against different state.

```{r accept}
if (plan@valid) {
  graft_commit(store, plan)
}
```

Two habits make this durable in practice:

- Give each agent a distinct `producer` and a per-run `run_id`. Every accepted
  revision then answers "which agent, in which run, proposed this" through
  `graft_history()`.
- Set an `idempotency_key` per proposed batch. A retried run — the common
  failure mode for agents — replays instead of writing a duplicate revision.

`graft_ingest()` collapses plan and commit into one call. Use it for a trusted
pipeline; keep the two-step form when an agent is the producer and something,
or someone, should look at the change first.

```{r history}
graft_history(store, "person:clark-kent")[
  , c("revision_number", "committed_at", "producer", "changed_fields")
]
```

## Give a file-editing agent a working surface

Coding agents are good at reading and editing files and less good at operating
a database client. Opening a store with `okf = "managed"` maintains an Open
Knowledge Format working tree: a deterministic Markdown projection of accepted
knowledge that an agent can read, grep, and edit directly. A file-backed store
uses a sibling directory, so `team.duckdb` projects into `team.okf` unless
`okf_path` says otherwise.

```{r okf}
okf_store <- graft_open(schema, "team.duckdb", okf = "managed")
graft_sync(okf_store)
graft_status(okf_store)
```

An edited file is a proposal, not accepted knowledge. `graft_status()` reports
that the tree is modified, and `graft_review()` turns the edits into the same
plan type `graft_plan()` returns:

```{r review}
edit_plan <- graft_review(
  okf_store,
  provenance = graft_provenance(
    producer = "directory-agent",
    run_id = "run-8843"
  )
)

edit_plan@changes[, c("class", "record_id", "action", "changed_fields")]

if (edit_plan@valid) {
  graft_commit(okf_store, edit_plan)
  graft_sync(okf_store)
}
```

The loop is explicit at every step:

```text
accepted revisions -> sync -> readable files -> agent edits -> review -> commit -> sync
```

An agent with filesystem access can therefore work in the surface it handles
best without gaining a second, unreviewed route into accepted knowledge. See
[Work with open knowledge](open-knowledge-format.html) for the full working
tree.

## What the tools cannot do

The boundary is worth stating plainly, because it is what makes a model-driven
read safe to enable:

- No SQL, connection object, filesystem path, or network argument is reachable
  from a tool definition.
- No tool writes. Acceptance happens only through `graft_commit()`, in R, from
  a plan someone or some policy approved.
- Graph traversal is bounded in hops, nodes, and edges; every tabular result is
  bounded by an explicit limit.
- Only manifest-declared public fields are searchable and returnable. Fields
  the contract marks sensitive — `display: restricted` in data-dict — are not
  exposed by a tool because they are dropped from the underlying read.
- The host decides which provider receives the tools and remains responsible
  for tool authorization.

```{r close}
graft_close(store)
```

Read [Retrieve accepted knowledge](retrieval.html) for the read operations
behind the tools, [Review knowledge changes](knowledge-change-control.html) for
plans and commit preconditions, and [Architecture](architecture.html) for how
the revision ledger and its derived projections fit together.
