---
title: "Verify a replacement artifact store"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Verify a replacement artifact store}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Graft can preview exclusions and verify the exact objects retained in a separate
store. This supplies one part of a permanent Forget workflow. The application
must still authorize Forget, retire the old generation, prevent old backup
admission, and arrange disposal. The example uses synthetic content and leaves
the source unchanged.

## Preview the full effect

A correction does not remove the earlier revision. Exclusion starts with exact
references supplied by the host and follows every declared dependency backwards:
reports derived from the excluded evidence must also be excluded. To forget a
logical record completely, the host supplies every revision covered by that
authorization; selecting one revision does not select other revisions with the
same artifact ID.

```{r}
library(graft)
source_path <- tempfile("source-")
target_path <- tempfile("replacement-")
source <- graft_artifact_store(source_path, create = TRUE)
private <- graft_artifact_save(
  source, "note", charToRaw("Synthetic private observation"), "text/plain"
)
report <- graft_artifact_save(
  source, "report", charToRaw("Interpretation of the observation"), "text/plain",
  dependencies = list(private)
)
survivor <- graft_artifact_save(
  source, "note", charToRaw("Independent corrected observation"), "text/plain"
)
selected <- graft_artifact_select(source, list(report))
accepted <- graft_artifact_decide(
  source, "review", "review-1", expected = NULL,
  selection = selected, action = "accept", actor = "synthetic-reviewer",
  reason = "Example review", purpose = "research"
)
plan <- graft_artifact_replacement_plan(source, forget = list(private))
vapply(plan$removed$artifacts, function(ref) ref$id, character(1))
plan$removed$selections
plan$removed$streams
```

The plan excludes the private revision, its report, the selection, and the entire
review stream. If that stream had later accepted a correction, it would still be
excluded: its earlier records retain the forgotten selection. Independent streams
keep their original history. Fresh acceptance of surviving material from an
excluded stream requires a new host review.

A dependency must be declared when the artifact is saved. Graft cannot discover
private passages copied into unrelated payloads, titles, IDs, or decision reasons.
The host supplies those additional artifact roots and `forget_streams`. Shared
content bytes remain if a surviving revision still uses them. Vocabulary releases
follow the same exact dependency rules, including dictionary sources and literal
context.

## Copy and verify survivors

The target must be empty. An edited plan or a changed source fails before copying.
After copying, Graft verifies every target object and rechecks the source.

```{r}
target <- graft_artifact_store(target_path, create = TRUE)
verified <- graft_artifact_replace(source, target, plan)
identical(verified, plan$target)
rawToChar(graft_artifact_read(target, survivor)$bytes)
# The source is deliberately unchanged by replacement.
rawToChar(graft_artifact_read(source, private)$bytes)
```

Excluded references and selections cannot be read from the replacement. The
application must also reject references to the retired generation; it must not
silently resolve an old reference against the replacement, even if its content
survived. Graft's low-level reads do not check generation admission.

On failure, a partial target may remain. Keep it quarantined and retry with a
fresh empty store. These functions never delete either store or make a candidate
visible to application readers. With PostgreSQL, commit the host transaction
successfully before treating a verified candidate as retained.

## Inspect a complete logical image

```{r}
manifest <- graft_artifact_manifest(target)
manifest$format
length(manifest$objects)
identical(manifest$id, verified$id)
unlink(c(source_path, target_path), recursive = TRUE)
```

A manifest records each object's kind, storage key, byte count, and digest. It
includes historical revisions, selections, and decisions, not just current
accepted heads. Well-formed orphan content is inventoried but omitted from a
replacement. Unknown entries, staging files, symbolic links, corrupt bytes,
incomplete dependencies, or invalid history prevent certification.

The manifest and plan can contain private identifiers and linkable hashes. Treat
them as restricted operational material, not public audit logs. Their digests
provide integrity; they provide neither authenticity nor freshness.

The defaults bound the inventory to 10,000 objects and 64 MiB of total stored
bytes. `max_metadata_bytes` bounds selection metadata and each complete decision
journal; artifact payload and revision limits also come from the store handle.
The host must quiesce writers for the entire operation. Local files retain their
trusted, single-writer model. PostgreSQL scope locks operate inside host-owned
transactions; copying between stores is not a distributed transaction.

## Keep restore admission outside the backup

Before replacing an affected Reader generation, the application must durably
retire it in an independent Forget journal. That journal must remain available
when an older content backup is restored. The host checks current admission on
reads, background jobs, cache delivery, imports, and restore; missing or stale
journal state denies access. A checksum-matching old backup remains retired.

The application also inventories external exports, provider context, Commons
working files, indexes, and reuse checkpoints. Deleting a Conversation does not
silently delete a separately accepted Memory or Artifact. Each requires the
appropriate preview and authorization.

Generation retirement prevents application resurrection only when every access
path enforces it. It does not destroy offline copies, invalidate available
encryption keys, or prove physical secure erasure. Graft's synthetic replacement
and restore tests do not establish those production guarantees.

The [Forget and restore gate](https://github.com/JamesHWade/graft/issues/48)
remains open until Rill's actual workflow, independent durable journal, closed
backups, and failure recovery are verified.

The next implementation tasks are [Rill Reader Forget](https://github.com/JamesHWade/rill/issues/106)
and [closed backups and publication recovery](https://github.com/JamesHWade/graft/issues/86).
