---
title: "Verify a replacement artifact store"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Verify a replacement artifact store}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Graft can preview exclusions and verify the exact objects retained in a separate
store. This is one step in a permanent Forget workflow. The application must
still authorize Forget, retire the old generation, prevent old backups from being
admitted, and arrange disposal. The example uses synthetic content and leaves the
source unchanged.

## Preview the full effect

A correction leaves the earlier revision in place. Exclusion begins with exact
references from the application and follows every declared dependency backwards, so
reports derived from excluded evidence are excluded too. To forget a logical
record completely, the application must supply every revision covered by the
authorization. Selecting one revision does not select other revisions with the
same artifact ID.

```{r}
library(graft)
source_path <- tempfile("source-")
target_path <- tempfile("replacement-")
source <- graft_store(source_path, create = TRUE)
private <- graft_save(
  source, "Synthetic private observation", id = "note", media_type = "text/plain"
)
report <- graft_save(
  source, "Interpretation of the observation", id = "report",
  media_type = "text/plain", dependencies = private
)
survivor <- graft_save(
  source, "Independent corrected observation", id = "note", media_type = "text/plain"
)
accepted <- graft_accept(
  source, report, stream = "review", key = "review-1", expected = NULL,
  actor = "synthetic-reviewer",
  reason = "Example review", purpose = "research"
)
plan <- graft_plan_replacement(source, forget = list(private))
vapply(plan$removed$artifacts, function(ref) ref$id, character(1))
plan$removed$selections
plan$removed$streams
```

The plan excludes the private revision, its report, the selection, and the entire
review stream. The stream remains excluded even if it later accepted a correction,
because its earlier records retain the forgotten selection. Independent streams
keep their original history. A fresh acceptance of surviving material from an
excluded stream requires a new application review.

Declare dependencies when saving an artifact. Graft cannot discover private
passages copied into unrelated payloads, titles, IDs, or decision reasons. The
application supplies any additional artifact roots and `forget_streams`. Shared content
bytes remain when a surviving revision still uses them. Vocabulary releases
follow the same exact dependency rules, including dictionary sources and literal
context.

## Copy and verify the remaining artifacts

The target must be empty. Graft rejects an edited plan or changed source before
copying. After copying, it verifies every target object and rechecks the source.

```{r}
target <- graft_store(target_path, create = TRUE)
verified <- graft_replace(source, target, plan)
identical(verified, plan$target)
graft_read(target, survivor)@data
# The source is deliberately unchanged by replacement.
graft_read(source, private)@data
```

Excluded references and selections cannot be read from the replacement. The
application must reject references to the retired generation. It must not
silently resolve an old reference against the replacement, even when its content
survived. Graft's low-level reads do not check whether the generation is admitted.

A partial target may remain after failure. Keep it quarantined and retry with a
fresh empty store. These functions never delete either store or expose a
candidate to application readers. With PostgreSQL, commit the application-owned transaction
successfully before treating a verified candidate as retained.

## Inspect a complete logical image

```{r}
manifest <- graft_manifest(target)
manifest$format
length(manifest$objects)
identical(manifest$id, verified$id)
unlink(c(source_path, target_path), recursive = TRUE)
```

A manifest records each object's kind, storage key, byte count, and digest. It
includes historical revisions, selections, and decisions, rather than only
current accepted heads. Well-formed orphan content is inventoried but omitted
from a replacement. Unknown entries, staging files, symbolic links, corrupt
bytes, incomplete dependencies, or invalid history prevent certification.

The manifest and plan can contain private identifiers and linkable hashes. Keep
them as restricted operational material rather than public audit logs. Their
digests provide integrity; they provide neither authenticity nor freshness.

By default, the inventory is limited to 10,000 objects and 64 MiB of stored
bytes. `max_metadata_bytes` limits selection metadata and each complete decision
journal. Artifact payload and revision limits also come from the store handle.
A local store's manifests, plans, and copies hold its store lock exclusively, so
writers in other processes wait for them rather than changing the store midway.
An application switching to a replacement holds the source's lock across the plan,
the copy, and the switch with `graft_with_store_lock()`, and checks inside each of
its own writes that the store is still current. PostgreSQL scope locks run inside
application-owned transactions; copying between stores is not a distributed
transaction.

## Clean up after failed saves

A failed save can leave content that no revision references. A plan with nothing
to forget copies everything else:

```{r, eval = FALSE}
cleanup <- graft_plan_replacement(store)
graft_replace(store, graft_store(fresh_path, create = TRUE), cleanup)
```

The application then switches to the fresh store as it would after a Forget.

## Keep the restore decision outside the backup

Before replacing an affected Reader generation, the application must retire it
durably in an independent Forget journal. The journal must remain available when
an older content backup is restored. The application checks whether access is
currently allowed for reads, background jobs, cache delivery, imports, and
restore. Missing or stale journal
state denies access. A checksum-matching old backup remains retired.

The application also inventories external exports, provider context, Commons
working files, indexes, and reuse checkpoints. Deleting a Conversation does not
silently delete a separately accepted Memory or Artifact. Each requires the
appropriate preview and authorization.

Generation retirement prevents an application from resurrecting content only when
every access path enforces it. It does not destroy offline copies, invalidate
available encryption keys, or prove physical secure erasure. Graft's synthetic
replacement and restore tests do not establish those production guarantees.

The [Forget and restore gate](https://github.com/JamesHWade/graft/issues/48)
remains open until Rill's actual workflow, independent durable journal, closed
backups, and failure recovery are verified.

Next implementation tasks are [Rill Reader Forget](https://github.com/JamesHWade/rill/issues/106)
and [publication and journal recovery](https://github.com/JamesHWade/graft/issues/86).
Graft now supplies [closed backup bundles and strict restore verification](artifact-backups.html).
These mechanics do not complete the application recovery protocol.
