---
title: "Supported integrations"
description: >
  Choose supported ellmer, data-dict, and Commons versions and reproduce
  the checks behind Graft's integration contracts.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supported integrations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Graft's store, plan, commit, and retrieval functions work without an agent
package. Add the dependencies for the interfaces you use:

| Interface | Tested baseline |
|---|---|
| Resolved data-dict JSON | Export format `0.1.0` |
| YAML authoring | CLI `0.0.1`, revision [d794c96](https://github.com/tidyverse/data-dict/tree/d794c9616f7803199432e9b31b519216aa78d1b0) |
| Agent tools and verification | ellmer **0.5.0** (also the required minimum) |
| Detached Commons source | Current upstream Commons R package, with ellmer 0.5.0 |

Compiled contracts and resolved JSON run with Graft's R dependencies alone.
They do not need a CLI, Python interpreter, ellmer, or Commons. Install the
data-dict CLI only where you author contracts. Agent tools can be invoked locally
and recorded answers verified without credentials or network requests. Commons
is optional and owns its source, agent loop, and labels.

Commons is not pinned to a version or revision. CI installs its current upstream
R package and exercises the detached-source contract. At runtime, Graft checks
that Commons exports `data_source()` with the arguments it needs; additional
arguments are allowed. These checks diagnose incompatible constructor changes,
while integration tests check behavior.

[Reuse narrative knowledge](ecosystem.html) gives executable recipes for current
ellmer, Deputy and dsprrr public interfaces. The host compatibility suite covers
real sync/streaming tool loops with a loopback-only synthetic server. Full
Tempest/Rill application reuse and scans/vitals evaluation remain in the
[integration backlog](https://github.com/JamesHWade/graft/issues/34).

The 2026-09-05 proof used ellmer 0.5.0, Deputy
`14e378951bb0f36a2a76fad0eff69b3cf8acab38`, and dsprrr
`76a014f74e8d4a236aa0a0c5dda87de388310cf6`. These record test evidence;
Graft's optional host dependencies are not revision-pinned. Run
`Rscript tools/check-host-composition.R` to check installed versions; missing
hosts fail that runner rather than silently skipping integration coverage.

Two real data-dict producers are compared separately: baseline 0.0.1 at
`d794c9616f7803199432e9b31b519216aa78d1b0`, and release 0.0.3 at
`9b48e97a61c4d12600cbb9dff3b1bb2aa14f6f71`. Both export the narrative fixture and
reject unknown root extensions. Both can export nested types that Graft's narrow
profile rejects. Release 0.0.3 reports the deliberately violated assertion;
the baseline does not. Graft itself treats those assertions as metadata.
Release 0.0.3 also renders descriptions as HTML and adds expression translations,
so it has its own frozen resolved fixture rather than an assumed byte-identical
export. Current upstream documentation is not evidence that the baseline CLI
implements every described feature.

## Use ellmer 0.5.0 or later

```{r install-ellmer}
install.packages("ellmer")
packageVersion("ellmer")
```

Both `graft_tools()` and `graft_verify()` check this minimum at runtime and give
an installation error for an older ellmer. The package uses public tool
definitions, `AssistantTurn`, `UserTurn`, tool request/result content, and
`AssistantPartialTurn`. Completed answers receive one verification row;
interrupted partial answers do not.

The old declaration allowed ellmer 0.3.0, whose recorded turns could be silently
omitted by the verifier. Graft now requires the 0.5.0 baseline rather than
maintaining multiple transcript representations. See ellmer's
[release notes](https://ellmer.tidyverse.org/news/index.html) for its changes.
Follow [Work with agents](agents.html) to register tools, pin accepted knowledge,
and inspect verification. Verification classifies the recorded evidence path;
it does not fact-check an answer or authenticate receipt identifiers.

## Reproduce the data-dict artifacts

The [data-dict specification](https://data-dict.tidyverse.org/spec.html) describes
the source format. Graft implements the narrower profile documented in
[Use a data-dict contract](data-dict-schema.html). Unsupported export versions
and unsupported shapes produce errors instead of silent conversions.

Build the tested CLI in an isolated directory:

```bash
cargo install --locked \
  --git https://github.com/tidyverse/data-dict \
  --rev d794c9616f7803199432e9b31b519216aa78d1b0 \
  --root /tmp/graft-data-dict \
  data-dict-cli
```

From the repository root, compare the real producer with all three committed
exports and compare the resulting structural contract digests:

```bash
GRAFT_TEST_DATA_DICT_CLI=/tmp/graft-data-dict/bin/data-dict \
  Rscript -e 'devtools::test(filter = "^data-dict-cli$", stop_on_failure = TRUE)'
```

The ordinary suite skips this producer check unless the environment variable
is set. Other data-dict tests use local exports and fake CLI calls; they never
install a compiler or access a provider.

When intentionally refreshing fixtures, export each source with the selected
producer and review the JSON diff before replacing the committed file:

```bash
/tmp/graft-data-dict/bin/data-dict export-spec \
  inst/extdata/team-directory.data-dict.yaml --pretty \
  > /tmp/team-directory.data-dict.json
diff -u inst/extdata/team-directory.data-dict.json /tmp/team-directory.data-dict.json
```

Repeat for `tests/testthat/fixtures/data-dict/personinfo/data-dict.yaml` and
`tests/testthat/fixtures/data-dict/tempest/data-dict.yaml`, comparing each with
its sibling `data-dict.export.json`. The parity test compares parsed exports,
so formatting-only differences do not count as semantic changes. When changing
the producer, update the recorded revision and expected CLI/export versions
together with the fixture evidence and compatibility workflow.

The CLI's `--version` does not establish its Git revision. Graft records the
executable digest and observed version; a configured revision is caller-supplied
provenance. See [Contract compiler details](contract-compilers.html). If Graft's
adapter implementation changes, regenerate and check
`inst/schema/graft-data-dict-adapter.source.json` with
`tools/update-data-dict-adapter-source.R` and update the manifest schema's pinned
adapter digest.

## Use the current Commons integration

```{r install-commons}
pak::pak("posit-dev/commons/pkg-r")
```

`graft_commons_data_source()` copies selected public accepted tables at one
boundary into a detached source. It retains typed empty tables, omits restricted
columns, and exports accepted definitions. Later Graft commits do not change
the copy; create a new source to pick up a newer boundary.

Commons keeps its own file measures, fallback queries, and evidence labels.
Its conversations do not acquire Graft verification merely because their data
came from this adapter. See the
[agent guide](agents.html#use-the-same-boundary-with-commons) for composition.

## Run the compatibility checks

The `compatibility.yaml` workflow installs dependencies before its offline checks:

- Older ellmer (0.4.0) must be rejected explicitly, including a chat containing
  an actual recorded assistant answer.
- ellmer 0.5.0 exercises bounded tools and verification through public ellmer
  objects, including a real Graft calculation receipt and interrupted turns.
  It also exercises dictionary discovery and structured proposals through a
  real `chat_structured()` call with local fake HTTP responses.
- The pinned data-dict CLI reproduces the exports; current upstream Commons constructs
  real detached sources with historical, restricted-field, and typed-empty cases.

The normal R package check resolves the available ellmer release satisfying
`>= 0.5.0`, while the separate baseline job keeps testing exactly 0.5.0. A future
release failure calls for reviewing the changed public interface, not relaxing
receipt or validation checks.

To reproduce the ellmer check, install the chosen version into a separate R
library and start a fresh process:

```bash
R_LIBS=/path/to/isolated-library \
  Rscript tools/check-ellmer-compatibility.R 0.5.0
```

The script asserts the loaded version and prints its library path, preventing
an installed development build from silently standing in for the test version.
Dependency installation is separate from the check. Tests create local stores
and recorded chats and never call a model.
