---
title: "Build a governed local Coworker"
description: >
  Combine Graft, Tempest, dsprrr, shinychat, and extensible ellmer tools in a
  local agent that publishes and remembers work only after approval.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a governed local Coworker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

An agent demo can look impressive while leaving the important product
questions unanswered. Which evidence may it use? Which tools may it call?
What happens before a side effect? Which results become durable context for
the next request?

Graft Coworker is an installable local work-surface example inspired by
[OpenWorker](https://github.com/andrewyng/openworker). It shows one answer for
R applications:

```text
request -> plan -> reconcile evidence -> prepare deliverable
        -> await approval -> publish file -> remember accepted outcome
```

The example is not a port of OpenWorker. It preserves the useful product
contract---outcome-oriented work, a reviewable action, extensible tools, and
continuity---while assigning each responsibility to an R package that already
has the right abstraction.

## Where each package fits

| Component | Responsibility |
|---|---|
| **Shiny and bslib** | Multi-user-safe application shell, responsive work view, approval inbox, memory ledger, and activity view |
| **shinychat and ellmer** | Conversation UI, model clients, and per-session tool loop |
| **dsprrr** | One planner contract with deterministic and model-backed implementations |
| **Tempest** | Typed workflow, artifacts, ordered events, and the approval pause before publication |
| **Graft** | Schema-checked storage for approved runs, deliverables, decisions, source links, and workspace memory |

The host application owns scheduling, routing, credentials, and side effects.
Graft remains a workflow-generic knowledge layer; the example does not add
Coworker-specific functions to the package API.

## Run the complete loop without a model key

From an installed package, locate and launch the example:

```{r run-installed, eval = FALSE}
library(graft)

coworker <- system.file(
  "examples",
  "coworker",
  package = "graft",
  mustWork = TRUE
)

shiny::runApp(file.path(coworker, "app"))
```

Choose **Run provider-free example**. The deterministic dsprrr planner uses a
bounded synthetic workspace containing GitHub, Jira, and Slack snapshots. It
prepares a release-readiness brief, then stops before publishing the Markdown
file.

The four main views expose different parts of the contract:

- **Work** shows the request, plan, source context, and finished deliverable.
- **Inbox** shows the exact proposed action and file target.
- **Memory** contains only outcomes accepted into Graft.
- **Activity** shows ordered Tempest workflow and approval events.

The provider-free path is useful for testing the complete lifecycle. It uses
the same planner interface and workflow as a model-backed run, so replacing
the planner does not change the approval or storage boundary.

From a development checkout, load the current Tempest source before Graft:

```{r run-checkout, eval = FALSE}
devtools::load_all("../tempest")
devtools::load_all(".")
shiny::runApp("inst/examples/coworker/app")
```

## Treat approval as a data boundary

Coworker does not use approval as a decorative confirmation dialog. Tempest
pauses the workflow before its exporter runs. Graft remains unchanged until
the operator decides what to do with the exact reviewed artifact.

| Decision state | Local file | Graft records and memory |
|---|---|---|
| Awaiting approval | Not written | Not written |
| Rejected | Not written | Not written |
| Approved | Reviewed Markdown is written | Run, deliverable, decision, source links, and memory are committed atomically |

Approval resumes the same Tempest run; it does not ask the planner to generate
the deliverable again. The path shown in the Inbox comes from the same
application helper used by the exporter, so the operator sees the file that
will actually be written.

The next request receives a bounded, deterministic view of accepted workspace
memory. Drafts and rejected outcomes never become continuity context.

This separation is the main reason to combine Tempest and Graft:

- Tempest controls what may happen next in a running workflow.
- Graft controls what accepted knowledge becomes durable and retrievable.

## Configure models through Tempest

The easiest model configuration remains Tempest's ambient option:

```{r ambient-model, eval = FALSE}
options(tempest.chat = "openai/gpt-5-mini")
shiny::runApp(file.path(coworker, "app"))
```

`tempest.chat` may also be an ellmer `Chat` object. Coworker deep-clones the
configured client for each browser session and uses a separate clone for the
planner, preserving the original template and its provider instructions.

Use a complete Tempest configuration when the assistant and planner need
different models:

```{r role-models, eval = FALSE}
options(
  graft.coworker.tempest_config = tempest::tempest_config(
    models = list(
      coordinator = "anthropic/claude-sonnet-4-20250514",
      writer = "openai/gpt-5-mini"
    )
  )
)
```

The Coworker assistant maps to Tempest's `coordinator` role and the dsprrr
planner maps to `writer`. Tempest's normal precedence rules still apply: an
explicit configuration wins over the ambient `tempest.chat` value, and an
explicit `chat_fn` wins over model strings and templates. The `chat_fn` escape
hatch supports an internal provider or any other custom ellmer client without
changing the workflow.

## Extend the tool registry

Coworker always registers two governed core tools:

- `prepare_outcome` starts the typed workflow and stops at approval; and
- `recall_memory` performs bounded retrieval over accepted Graft memory.

Add ordinary ellmer tools with an option:

```{r add-tool, eval = FALSE}
installed_package_version <- ellmer::tool(
  function(package) as.character(utils::packageVersion(package)),
  "Return the installed version of an R package.",
  arguments = list(
    package = ellmer::type_string("An installed R package name.")
  ),
  name = "installed_package_version"
)

options(
  graft.coworker.tools = list(installed_package_version)
)
```

A provider function can construct session-aware tools. It receives the current
`worker`, `input`, `revision`, `session`, and `example_dir`:

```{r tool-provider, eval = FALSE}
options(
  graft.coworker.tools = function(context) {
    list(my_session_tool(context$session))
  }
)
```

Coworker rejects duplicate tool names instead of silently replacing a tool.
This makes it safe to compose several tool providers while keeping the final
registry inspectable.

### Add the Posit `btw` tool belt

The optional [posit-dev/btw](https://github.com/posit-dev/btw) integration can
discover a broad R-aware tool collection:

```{r btw-read-only, eval = FALSE}
install.packages("btw")
options(graft.coworker.btw = "read_only")
```

The `read_only` profile admits environment, documentation, file-reading, Git
inspection, session, skill, CRAN, IDE, and web-reading tools. It excludes file
edits, package execution, arbitrary R execution, Git mutations, GitHub API
access, and subagents.

Select btw groups or tool names for a smaller profile:

```{r btw-subset, eval = FALSE}
options(graft.coworker.btw = c("docs", "env", "sessioninfo"))
```

`options(graft.coworker.btw = "all")` exposes every locally available btw tool
and is appropriate only for a trusted single-user environment.

Directly registered mutating tools do not automatically pass through
Coworker's publication approval. A production write tool should be a typed
Tempest capability whose exact action and target can be shown before
execution.

## Replace the reference workspace

The example keeps application concerns under `inst/examples/coworker/`:

- `corpus/project-atlas.json` is the bounded reference source bundle;
- `schema/coworker.linkml.yaml` defines the application-owned record contract;
- `R/planner.R` defines the dsprrr planning interface;
- `R/workflow.R` defines artifacts, approval, and publication;
- `R/worker.R` owns the Graft store and approved ingestion; and
- `app/` contains the Shiny host.

A production host can replace the JSON bundle with authenticated read
adapters, add Tempest capabilities for external actions, and preserve the same
review boundary. Its record schema can evolve independently of Graft's package
API.

The default data directory is
`file.path(tools::R_user_dir("graft", "data"), "coworker")`. Set
`options(graft.coworker.data_dir = "/path")` or the
`GRAFT_COWORKER_HOME` environment variable before launch to use another local
directory.

## Know the current boundary

This is a vertical slice, not a desktop-agent replacement. It does not yet
provide a desktop shell, background scheduling, OAuth connector catalog,
durable chat transcripts, voice input, MCP client, or real Slack and calendar
writes.

Those are host features rather than missing Graft record functions. The
current example is deliberately complete at one narrower boundary: prepare a
finished local work product, review the exact publication, and make only the
accepted outcome durable.

Continue with [Getting started with Graft](getting-started.html) for the
schema-to-store workflow, or
[Govern knowledge changes](knowledge-change-control.html) for revision history
and reviewed schema evolution. The complete Coworker implementation is
available in the
[source directory](https://github.com/JamesHWade/graft/tree/main/inst/examples/coworker).
