---
title: "Use a data-dict contract with graft"
description: >
  Start from related tables, resolve their data dictionary, and understand
  exactly which rules graft enforces.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use a data-dict contract with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Use data-dict when the knowledge you have today is tabular. A data dictionary
names the tables and columns, documents their meaning, and declares constraints
and relationships. graft compiles the supported parts into the contract used
to review candidate records.

The source contract does not become the knowledge store. Accepted records and
their revisions live in graft's DuckDB ledger.

## Author once, then use the resolved dictionary

The package includes a small team directory in two forms:

- `team-directory.data-dict.yaml` is the human-authored source; and
- `team-directory.data-dict.json` is the resolved output from `data-dict
  export-spec`.

Contract authors install the
[data-dict CLI](https://data-dict.tidyverse.org/), validate the readable YAML,
and resolve it once:

```bash
data-dict validate-spec team-directory.data-dict.yaml
data-dict export-spec team-directory.data-dict.yaml --pretty > team-directory.data-dict.json
```

The rest of this guide starts from the shipped resolved JSON. That keeps the
runnable package lesson R-only; the bundled JSON is the exact export of the
bundled YAML under the CLI revision documented in the
[compiler reference](contract-compilers.html).

```{r compile-json}
library(graft)

resolved_path <- system.file(
  "extdata",
  "team-directory.data-dict.json",
  package = "graft",
  mustWork = TRUE
)

schema <- graft_schema(resolved_path)

schema@name
schema@version
names(schema@classes)
```

When compiling this resolved source without `output`, `graft_schema()` writes a
temporary `.graft.json` manifest. Supply an output path when the compiled
contract should be committed or deployed as an artifact:

```{r durable-output, eval = FALSE}
schema <- graft_schema(
  resolved_path,
  output = "team-directory.graft.json"
)
```

## See how tables become a contract

The source YAML uses familiar table concepts. Its essential structure is:

```yaml
$version: 0.1.0
$learn_more: https://data-dict.tidyverse.org/
name: team_directory
version:
  number: 0.1.0

tables:
  - name: organization
    columns:
      - name: id
        type: string
        constraints: [primary_key]
        examples: ["org:daily-planet"]
      - name: name
        type: string
        constraints: [required]
        examples: ["Daily Planet"]

  - name: person
    columns:
      - name: id
        type: string
        constraints: [primary_key]
        examples: ["person:lois-lane"]
      - name: full_name
        type: string
        constraints: [required]
        examples: ["Lois Lane"]
      - name: job_title
        type: string
        examples: ["Reporter"]

  - name: employment
    columns:
      - name: id
        type: string
        constraints: [primary_key]
        examples: ["employment:lois-lane:daily-planet"]
      - name: person_id
        type: string
        constraints: [required, foreign_key]
        examples: ["person:lois-lane"]
      - name: organization_id
        type: string
        constraints: [required, foreign_key]
        examples: ["org:daily-planet"]

relationships:
  - join: employment.person_id = person.id
    cardinality: many-to-one
  - join: employment.organization_id = organization.id
    cardinality: many-to-one
```

The adapter maps each table to a Graft class and each column to a slot. Primary
keys establish record identity. Scalar foreign keys become reference checks.

```{r inspect}
employment <- schema@classes$employment

names(employment@slots)
employment@slots$person_id@required
employment@slots$person_id@object_reference
employment@slots$person_id@range
```

Candidate inputs use the same table names:

```{r candidate-shape}
records <- list(
  organization = data.frame(
    id = "org:daily-planet",
    name = "Daily Planet"
  ),
  person = data.frame(
    id = "person:lois-lane",
    full_name = "Lois Lane",
    job_title = "Reporter"
  ),
  employment = data.frame(
    id = "employment:lois-lane:daily-planet",
    person_id = "person:lois-lane",
    organization_id = "org:daily-planet"
  )
)
```

The [getting-started guide](getting-started.html) opens an empty store, shows a
missing foreign key being rejected, commits this corrected candidate set, and
retrieves two revisions of the person record.

## Know what the table profile enforces

The current `graft-table-v1` profile is deliberately narrower than data-dict
itself:

- every table has exactly one primary key named `id`, represented as a scalar
  string;
- IDs are non-empty and globally unique across all tables in one Graft store;
- every YAML column declares a type;
- non-key columns are scalars or one-level lists of scalar values;
- foreign keys are scalar strings that target another table's primary `id`;
  and
- resolved `foreign_key` constraints and `references` metadata must agree.

Struct columns, nested lists, list-valued foreign keys, and numeric ID types are
rejected rather than converted. Use quoted strings for identifiers and codes.

Some data-dict metadata remains descriptive. Graft does not currently execute
non-primary uniqueness, representative ranges, assertion text, relationship
cardinality, join aliases, conflicts, range joins, or multi-column joins.
Scalar foreign keys are validated references, not graph traversal edges.

One column key does change what reads return. A column marked
`display: restricted` is compiled as sensitive: its values are dropped from
retrieved records, excluded from searchable fields, and never reach a tool
built with `graft_tools()`. Primary IDs are public operational keys and cannot
be restricted. This is the contract-level control over what an agent can see;
see [Work with agents](agents.html).

The compiled mapping report records these boundaries:

```{r mapping-report}
schema@manifest$dictionary$profile
unlist(schema@manifest$dictionary$mapped[c(
  "tables_to_classes",
  "columns_to_slots",
  "foreign_keys_to_object_references"
)])
```

The [compiler reference](contract-compilers.html) is the authoritative list of
descriptive or unsupported data-dict semantics.

## Compile YAML when authoring the contract

The YAML route lets data-dict resolve its own source format. It requires the
optional `data-dict` executable; graft never downloads or installs it during
package use.

```{r compile-yaml, eval = FALSE}
yaml_path <- system.file(
  "extdata",
  "team-directory.data-dict.yaml",
  package = "graft",
  mustWork = TRUE
)

options(graft.data_dict_cli = "/path/to/data-dict")

schema <- graft_schema(
  yaml_path,
  output = "team-directory.graft.json"
)
```

graft runs `data-dict export-spec`; it does not run data-dict's metadata or
data validation commands. Teams commonly resolve YAML in a schema-authoring or
CI environment and commit the YAML, resolved JSON, and compiled `.graft.json`
together.

Resolved JSON is trusted build input. graft checks its supported export version
and structure, but cannot prove that another program actually produced it with
`export-spec`.

See [Contract compiler details](contract-compilers.html) for CLI pinning,
compiler provenance, digest behavior, public-manifest redaction, exact numeric
and datetime rules, and the LinkML compiler's strict profile.

## Where to go next

Continue with [change control](knowledge-change-control.html) to review and
accept plans, then [retrieval and history](retrieval.html) to read the accepted
knowledge and its revisions.

Use LinkML when the contract needs inheritance, class or predicate URIs,
polymorphic references, Graft record roles, narrative or semantic statements,
custom identity policies, qualifier fields, or graph traversal relationships.
The [LinkML guide](linkml-schema.html) continues from the same plan and commit
model; only the contract provider changes.

## Discover the accepted dictionary

Use `graft_dictionary(store, table = "person", field = "full_name")` to inspect
public meaning and enforced contract properties with deterministic bounds and
an accepted-boundary receipt. The same operation is available to ellmer through
`graft_tools()`. See [Discover meaning before reading records](agents.html#discover-meaning-before-reading-records)
for selection, pagination, privacy, and verification behavior.
