---
title: "Reuse narrative knowledge with R agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reuse narrative knowledge with R agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(graft)
```

A research conclusion, reading interpretation, preference, or unresolved question
can be ordinary record content. data-dict describes its fields and relationships;
Graft adds reviewed acceptance, history and pinned reads. The application decides
what the content means and who may use it.

## Try the narrative workflow offline

This teaching example uses synthetic records and the shipped resolved dictionary.
It requires neither a model nor the data-dict CLI. Source the example into a local
environment, then run it:

```{r narrative-example}
example <- new.env()
sys.source(system.file("examples/narrative-knowledge.R", package = "graft"), example)
result <- example$narrative_example()
result$rejected[, c("field", "message")]
result$pinned$body
result$current$body
result$history[, c("revision_number", "producer")]
```

The example rejects a missing evidence target, accepts corrected records, pins a
reading, and accepts a later correction. The earlier reading retains its original
meaning. A preference has no fabricated evidence link; its purpose is reading
selection. Multiple evidence anchors are normalized support records.

The fixture is not Tempest's promotion schema or a Rill integration. Applications
keep their richer contracts. A stored `archived` value remains readable through
ordinary Graft APIs; it does not enforce automatic-consultation eligibility.
Likewise, restricted fields are omitted from public reads, but an `owner_binding`
column does not establish Reader authorization.

## Compose public tools with a host

Create one pinned tool set after the host has authorized the store and selection:

```{r composition-store}
store <- graft_open(graft_schema(system.file(
  "extdata/narrative-knowledge.data-dict.json", package = "graft"
)), okf = "disabled")
graft_ingest(store, example$narrative_records(),
  graft_provenance("example-host", idempotency_key = "example-1"))
view <- graft_at(store, graft_snapshot(store))
tools <- graft_tools(view)
tools$graft_get("knowledge:preference")$result$record
```

A host supplies its own configured Chat, credentials, permissions, context policy
and execution lifetime. These are alternative recipes for that same `tools` list;
the model calls are optional and are not run when building this guide:

```{r host-recipes, eval=FALSE}
# ellmer
chat$set_tools(tools)
chat$chat("Read the accepted interpretation and quote its basis.")
graft_verify(chat)

# Deputy: retain canonical results in the transcript for this bounded recipe.
agent <- deputy::Agent$new(chat = chat, tools = tools,
  context_policy = deputy::ContextPolicy(
    max_tokens = NULL, max_tool_result_bytes = NULL
  ))
agent$run_sync("Read the accepted interpretation and quote its basis.")
graft_verify(chat)

# Current dsprrr uses the dedicated react() constructor.
module <- dsprrr::react(dsprrr::signature("question -> answer"),
  tools = tools, chat = chat)
answer <- dsprrr::run(module, question = "What does the interpretation say?")
graft_verify(chat)
```

Use a fresh Chat for each host. The tests execute these public paths with actual
objects and a loopback-only HTTP/SSE server: pinned reads after later acceptance,
Unicode/Markdown, row truncation, missing IDs, and mixed non-Graft evidence.
Verification consumes the actual ellmer Chat. Deputy run events or dsprrr's
summary metadata are not a substitute transcript. dsprrr's final structured
answer is not a completed text-bearing answer classified by `graft_verify()`;
inspect the recorded text answers it actually contains.

A tool name collision must be checked before combining host and Graft tools:

```{r collision-recipe, eval=FALSE}
combined <- c(tools, host_tools)
tool_names <- vapply(combined, function(tool) tool@name, character(1))
stopifnot(!anyDuplicated(tool_names))
chat$set_tools(combined)
```

Check the ToolDef name, not just its outer list name. ellmer replaces an existing
name with a message, dsprrr retains the supplied tools and registers them with
ellmer, and Deputy rejects duplicate registration. This host-side check avoids
relying on different collision behavior. It does not require a Graft adapter.

## Preserve results without overstating verification

ellmer 0.5 stores ordinary list tool results as JSON. Graft's verifier decodes
that envelope and applies the same receipt and citation checks; malformed JSON
and duplicate object keys fail closed. A cited narrative requires an explicit
matching quotation. Mixed non-Graft evidence or a failed tool cannot become
verified just because another tool returned a Graft receipt.

Graft currently returns ordinary R lists from its tools. ellmer 0.5 accepts them
but emits a deprecation warning for implicit JSON conversion. The host tests
record this compatibility limitation; [#53](https://github.com/JamesHWade/graft/issues/53)
owns the public return representation decision. Do not suppress all tool warnings
in application code.

Bounds have distinct meanings. Search/history and related-record limits bound
rows. Dictionary discovery additionally bounds text. `graft_get()` retains the
full public narrative field and does not offer text pagination or a character
ceiling. Size the authorized records and context budget accordingly.

Deputy's default context policy may offload a large result to a durable reference.
The host can retrieve the exact R value with `agent$resolve_tool_result(reference)`;
the model has a bounded `deputy_read_tool_result` path. Tests recover a long
Unicode result and its identical receipt. The offloaded reference/preview is not
the canonical Graft envelope, so `graft_verify()` reports that evidence path as
untrusted. Disabling offloading in the recipe above is an explicit choice for a
bounded example, not a general recommendation for large documents.

## Reconnect in the process that reads

A live DuckDB connection belongs to its process. Keep serializable, host-trusted
references and reconstruct a view in a worker. The regression test closes the
store before a fresh R process opens it read-only and verifies a complete selected
record/evidence set. Do not deserialize arbitrary untrusted RDS content.

```{r worker-recipe, eval=FALSE}
# Host-owned checkpoint: retain the full selection on unchanged days.
basis <- list(snapshot = graft_snapshot(store), ids = selected_ids,
  revisions = selected_revision_ids)
saveRDS(basis, checkpoint_path)
graft_close(store)

# In the worker: schema_path and store_path come from the authorized host.
basis <- readRDS(checkpoint_path)
store <- graft_open(graft_schema(schema_path), store_path,
  read_only = TRUE, okf = "disabled")
view <- graft_at(store, basis$snapshot)
records <- lapply(basis$ids, function(id) graft_get(view, id))
graft_close(store)
```

A snapshot identifies a boundary, not which records a task selected or permission
to access them. A full selected basis includes evidence and exact revisions;
new change receipts do not replace it. Eligibility, Reader scope, permanent
Forget and backup/restore remain explicit design/integration work.

## Ecosystem ownership and delivery status

| Package | Owns | Evidence in this slice |
|---|---|---|
| data-dict | Types, dictionary meaning, expressions | Real producer export and profile probes |
| Graft | Acceptance, identity, revisions, bounded reads, Definitions, Receipts | Offline narrative/correction/worker tests |
| ellmer | Chat, ToolDef, model transport | Real Chat/tool loop with synthetic HTTP responses |
| Deputy | Permissions, execution, offloading, host lifecycle | Real Agent loop and exact offloaded-result recovery |
| dsprrr | Modules, signatures, execution | Real `react()` loop and proposal signature |
| Commons | Detached public source, file measures, fallback labels | Existing detached-source compatibility suite; unpinned |
| Tempest | Scientific evidence, research products, promotion | Existing public contracts inventoried; full reuse proof in #50 |
| Rill | Readers, Documents, accepted reading outcomes, Archive/Forget | ADR-defined ownership; runtime proof and privacy gates in #47/#48/#51 |
| scans/vitals | Evaluation | Task-quality evaluation remains in #41 |

Acceptance is not factual truth, access permission or authority to execute saved
code. Graft Definitions retain their reviewed expression contract; Commons file
measures remain separate trusted functions. No companion schema is needed for
this teaching workflow. The [backlog](https://github.com/JamesHWade/graft/issues/34)
tracks the remaining application and lifecycle proofs separately.

## Reproduce compatibility evidence

From a checkout with the optional hosts installed:

```bash
Rscript tools/check-host-composition.R
GRAFT_TEST_DATA_DICT_CLI=/path/to/data-dict \
  Rscript -e 'devtools::test(filter = "^data-dict-narrative$", stop_on_failure = TRUE)'
```

The host runner prints installed package versions and available source revisions.
Tests do not install dependencies or contact model providers. See the
[compatibility guide](compatibility.html) for installation and the exact tested
producer versions, and [data-dict contracts](data-dict-schema.html) for the narrower
Graft adapter profile.

```{r close-store, include=FALSE}
graft_close(store)
```
