---
title: "Examples"
description: >
  See how ordinary LinkML schemas can describe chemical entities, biological
  samples, and the relationships between them.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

A useful schema does more than name columns. It distinguishes things that
should not be conflated and makes important relationships explicit.

This page contains two small examples:

- a chemistry model for a molecule, its stereoisomers, and a racemic mixture;
  and
- a biology model connecting a study, a biosample, and a sequencing process.

Both are ordinary LinkML schemas. Neither imports graft's core schema or uses
graft annotations. See [Use a LinkML schema with
graft](linkml-schema.html) for a slower introduction to compiling a schema.

The examples are deliberately smaller than the models that inspired them.
They preserve the useful modeling pattern without claiming conformance to the
full upstream standard.

## Chemistry: distinguish a molecule from its forms

[CHEMROF](https://chemkg.github.io/chemrof/home/) is a LinkML model for
chemical entities, mixtures, and related concepts. One of its examples models
thalidomide as two enantiomers and a racemic mixture.

That distinction matters. A name such as "thalidomide" is not enough to tell
us whether a record means the chirality-agnostic molecule, one stereoisomer,
or a mixture.

The example schema uses inheritance and typed references:

```yaml
classes:
  ChemicalEntity:
    abstract: true
    attributes:
      id:
        identifier: true
        required: true
        range: uriorcurie
      name:
        required: true

  Enantiomer:
    is_a: Molecule
    attributes:
      chirality:
        range: ChiralityEnum
        required: true
      chirality_agnostic_form:
        range: Molecule
        required: true
        inlined: false

  RacemicMixture:
    is_a: ChemicalEntity
    attributes:
      has_left_enantiomer:
        range: Enantiomer
        required: true
        inlined: false
      has_right_enantiomer:
        range: Enantiomer
        required: true
        inlined: false
```

The [complete chemistry
schema](https://github.com/JamesHWade/graft/blob/main/inst/extdata/chemistry.linkml.yaml)
also declares molecular formulae, aliases, semantic mappings, and an enum for
chirality.

Load the compiled manifest and inspect the concrete classes:

```{r chemistry-schema}
library(graft)

chemistry_schema <- kg_schema(system.file(
  "extdata",
  "chemistry.graft.json",
  package = "graft"
))

kg_classes(chemistry_schema)
kg_enums(chemistry_schema)
```

The data use CURIEs as their identifiers. References contain the identifier of
the target record.

```{r chemistry-data}
chemistry_records <- list(
  Molecule = data.frame(
    id = "chemistry:thalidomide",
    name = "thalidomide",
    molecular_formula = "C13H10N2O4"
  ),
  Enantiomer = data.frame(
    id = c("chemistry:S-thalidomide", "chemistry:R-thalidomide"),
    name = c("S-thalidomide", "R-thalidomide"),
    molecular_formula = c("C13H10N2O4", "C13H10N2O4"),
    chirality = c("left", "right"),
    chirality_agnostic_form = c(
      "chemistry:thalidomide",
      "chemistry:thalidomide"
    )
  ),
  RacemicMixture = data.frame(
    id = "CHEBI:9513",
    name = "racemic thalidomide",
    aliases = I(list(c("thalidomide", "thalidomide racemate"))),
    has_left_enantiomer = "chemistry:S-thalidomide",
    has_right_enantiomer = "chemistry:R-thalidomide",
    chirality_agnostic_form = "chemistry:thalidomide"
  )
)
```

Write all three classes in one batch. graft checks the references against both
the existing store and the other records staged in the batch.

```{r chemistry-ingest}
chemistry_store <- suppressMessages(
  kg_connect_duckdb(chemistry_schema, ":memory:")
)
kg_init(chemistry_store)

kg_ingest(
  chemistry_store,
  kg_batch(
    producer = "chemistry-example",
    idempotency_key = "thalidomide-v1"
  ),
  chemistry_records
)
```

The mixture can then be retrieved without flattening away the distinction
between the molecule and its stereoisomers:

```{r chemistry-query}
mixture <- kg_get(chemistry_store, "CHEBI:9513")

mixture$record[c("id", "name")]
mixture$record[c(
  "has_left_enantiomer",
  "has_right_enantiomer",
  "chirality_agnostic_form"
)]
```

This pattern also works for salts, isotopes, mixtures, reaction participants,
or any other case where chemistry-specific identity matters.

## Biology: connect samples to the work performed on them

The [National Microbiome Data Collaborative
schema](https://microbiomedata.github.io/nmdc-schema/) uses LinkML to describe
environmental omics data. Its model separates studies, biosamples, data
generation, workflow executions, and data objects.

The smaller schema here keeps three of those ideas:

```yaml
classes:
  Study:
    is_a: NamedThing

  Biosample:
    is_a: NamedThing
    attributes:
      associated_studies:
        range: Study
        required: true
        multivalued: true
        inlined: false
      ecosystem:
      collection_site:

  NucleotideSequencing:
    is_a: DataGeneration
    attributes:
      has_input:
        range: Biosample
        required: true
        multivalued: true
        inlined: false
      analyte_category:
        range: AnalyteCategoryEnum
        required: true
```

The [complete biology
schema](https://github.com/JamesHWade/graft/blob/main/inst/extdata/biology.linkml.yaml)
adds semantic mappings, coordinates, instrument details, and a controlled
vocabulary for the analyte.

```{r biology-schema}
biology_schema <- kg_schema(system.file(
  "extdata",
  "biology.graft.json",
  package = "graft"
))

kg_classes(biology_schema)
kg_slots(biology_schema, "Biosample")
```

The records make the provenance chain explicit: the sequencing process belongs
to a study and takes a particular biosample as input.

```{r biology-data}
biology_records <- list(
  Study = data.frame(
    id = "nmdc:sty-11",
    name = "Forest soil carbon study",
    objective = "Relate soil conditions to microbial carbon cycling.",
    ecosystem = "Environmental"
  ),
  Biosample = data.frame(
    id = "nmdc:bsm-11",
    name = "Forest soil core 11",
    associated_studies = I(list("nmdc:sty-11")),
    ecosystem = "Environmental",
    collection_site = "Example forest plot",
    latitude = 44.57,
    longitude = -85.61
  ),
  NucleotideSequencing = data.frame(
    id = "nmdc:dg-11",
    name = "Soil metagenome sequencing",
    associated_studies = I(list("nmdc:sty-11")),
    has_input = I(list("nmdc:bsm-11")),
    analyte_category = "metagenome",
    instrument_name = "Illumina NovaSeq",
    processing_institution = "Example sequencing center"
  )
)
```

```{r biology-ingest}
biology_store <- suppressMessages(
  kg_connect_duckdb(biology_schema, ":memory:")
)
kg_init(biology_store)

kg_ingest(
  biology_store,
  kg_batch(
    producer = "biology-example",
    idempotency_key = "soil-v1"
  ),
  biology_records
)
```

The tabular and hydrated interfaces answer different questions:

```{r biology-query}
kg_records(biology_store, "Biosample") |>
  dplyr::select(id, name, ecosystem, collection_site) |>
  dplyr::collect()

sequencing <- kg_get(biology_store, "nmdc:dg-11")
sequencing$record[c(
  "name",
  "analyte_category",
  "has_input",
  "associated_studies"
)]
```

The same pattern can connect specimens to assays, genes to phenotypes, or
organisms to observations. LinkML supplies the domain vocabulary and
constraints; graft supplies the local storage and retrieval layer.

## Use the upstream model when it fits

These examples are adaptations, not replacements for CHEMROF or NMDC. A
production project should first ask whether an established model already
describes its data.

Compile the upstream schema directly when it uses the LinkML constructs graft
supports and its classes match the records you need. Use a smaller schema when
you need only a well-defined subset, and retain upstream `class_uri`,
`slot_uri`, and identifier mappings so that the relationship remains visible.

The official [LinkML examples
page](https://linkml.io/linkml/examples.html) also points to Biolink, NMDC, and
other production models. The [LinkML Schema
Registry](https://linkml.io/linkml-registry/registry/) is useful when looking
for an existing model in another domain.

```{r cleanup, include = FALSE}
kg_disconnect(chemistry_store)
kg_disconnect(biology_store)
```
