---
title: "Getting started with graft"
description: >
  Follow one provider-free workflow from a LinkML contract through review,
  accepted history, retrieval, and open knowledge.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

graft gives an R workflow one explicit path from candidate records to accepted
knowledge:

1. Load a LinkML domain contract.
2. Describe where the candidate change came from.
3. Plan and review the change without writing.
4. Commit the reviewed plan atomically.
5. Retrieve current records and accepted history through bounded functions.
6. Synchronize a readable OKF working tree and review any edits through the
   same commit boundary.

The walkthrough uses only graft and package data. It does not require a model
provider or an orchestration framework.

## Load the contract

`graft_schema()` returns an immutable `GraftSchema`. It can load a compiled
`.graft.json` contract directly, or compile a LinkML YAML schema. Loading a
compiled contract does not require Python.

```{r contract}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "personinfo.graft.json",
  package = "graft"
))

schema@classes
schema@classes$Person@slots
```

The example contract requires explicit identifiers for `Organization` and
`Person`. A person's `employed_by` values must identify organizations. These
are domain rules, not storage details.

Open a file-backed store so that accepted revisions persist and graft can
manage a sibling OKF working tree:

```{r open}
store <- graft_open(schema, "knowledge.duckdb")
```

`graft_open()` initializes a new store or verifies an existing one against the
contract. The returned `GraftStore` owns the live connection state and can be
closed safely more than once with `graft_close()`.

## Describe provenance

Every candidate change has a `GraftProvenance` object. Give a stable
idempotency key when retrying the same producer event should return the
original result instead of creating another accepted observation.

```{r provenance}
provenance <- graft_provenance(
  producer = "directory-import",
  version = "1.2.0",
  run_id = "run-42",
  idempotency_key = "daily-planet-v1",
  metadata = list(source = "staff-directory.csv")
)
```

Provenance identifies the origin of the change. It is not a free-form substitute
for domain fields that belong in the records themselves.

## Plan before writing

Candidate records are a named list of data frames. The names are concrete
classes from the active contract. Use list-columns for multivalued fields.

```{r records}
records <- list(
  Organization = data.frame(
    id = "org:daily-planet",
    name = "Daily Planet"
  ),
  Person = data.frame(
    id = "person:clark-kent",
    full_name = "Clark Kent",
    aliases = I(list(c("Superman", "Kal-El"))),
    employed_by = I(list("org:daily-planet"))
  )
)
```

Planning normalizes the records, resolves identity, validates relationships,
and compares them with current accepted heads. It does not persist records or
provenance.

```{r plan}
plan <- graft_plan(store, records, provenance)

plan@valid
plan@changes
plan@issues
```

`plan@changes` is the review surface: it distinguishes inserts, updates, and
matches and identifies changed fields. `plan@issues` contains all validation
problems that could safely be collected in one pass, including source-row
coordinates. Invalid plans cannot be committed.

Planning also binds the proposal to the store, active schema, expected record
heads, and a deterministic digest. This makes a plan suitable for a human or
host policy to review before acceptance.

## Commit the reviewed plan

```{r commit}
if (plan@valid) {
  result <- graft_commit(store, plan)
}
```

Before writing, `graft_commit()` rechecks that the plan is intact and still
matches the store, contract, and record heads. It then accepts the complete
change atomically. A stale or altered plan fails without a partial commit.

When no separate review is needed, `graft_ingest()` is the convenience form:

```{r ingest}
result <- graft_ingest(store, records, provenance)
```

It is equivalent to planning and immediately committing a valid plan. It does
not define a second write path.

## Retrieve current knowledge

The retrieval surface is deliberately bounded. `graft_get()` hydrates one
current record, while `graft_find()` searches only public fields declared by
the contract.

```{r retrieval}
person <- graft_get(store, "person:clark-kent")
person$record

graft_find(
  store,
  query = "Clark",
  class = "Person",
  limit = 10
)
```

`graft_query()` collects advanced retrieval behind one validated interface
instead of exposing internal projections. The operation selects a fixed query
shape; the request supplies only operation-specific values.

```{r query}
graft_query(
  store,
  operation = "neighbors",
  request = list(
    id = "person:clark-kent",
    projection = "semantic",
    hops = 1,
    max_nodes = 25,
    max_edges = 50
  )
)

graft_query(
  store,
  operation = "integrity",
  request = list(projections = TRUE),
  limit = 100
)
```

Other operations cover exact identifier lookup, identifiers, claims,
evidence, bounded traversal, and unresolved mentions. Unknown request members
are rejected; arbitrary SQL is not part of the public contract.

## Inspect accepted history

Create a second plan when the source reports a change:

```{r update}
update <- list(
  Person = data.frame(
    id = "person:clark-kent",
    full_name = "Clark Kent",
    age = 35L,
    aliases = I(list(c("Superman", "Kal-El"))),
    employed_by = I(list("org:daily-planet"))
  )
)

update_plan <- graft_plan(
  store,
  update,
  graft_provenance(
    producer = "directory-import",
    run_id = "run-43",
    idempotency_key = "daily-planet-v2"
  )
)

update_plan@changes
graft_commit(store, update_plan)
```

`graft_history()` reads immutable revisions in deterministic newest-first
order. An optional accepted batch ID or timestamp selects an exact commit
boundary.

```{r history}
history <- graft_history(
  store,
  "person:clark-kent",
  limit = 20
)

history[, c("batch_id", "changed_fields", "record")]
```

Current retrieval and history are two views of the same accepted revision
ledger. Current records are not an independent write surface.

## Work through readable files

`graft_sync()` projects current accepted knowledge into the managed OKF working
tree. Synchronization never changes accepted records.

```{r okf-sync}
graft_sync(store)
graft_status(store)
```

People and agents can read and propose changes in Markdown and YAML. A local
edit is not accepted knowledge. `graft_review()` validates the working tree and
returns an ordinary `GraftCommitPlan`:

```{r okf-review}
review <- graft_review(
  store,
  provenance = graft_provenance(
    producer = "human-review",
    idempotency_key = "approved-okf-edit-1"
  )
)

review@changes
review@issues

if (review@valid) {
  graft_commit(store, review)
}
```

After acceptance, synchronize again to make the working tree reflect the new
head revisions.

`graft_tools()` creates read-only agent tools from the same bounded retrieval
operations. Creating the definitions does not require a provider:

```{r tools}
tools <- graft_tools(store)
names(tools)
```

The tools expose neither mutation nor raw database, filesystem, or network
access. The host chooses whether and how to attach them to a model.

## Close the store

```{r close}
graft_close(store)
graft_close(store)
```

Closing is idempotent. graft closes connections it created; a connection
supplied by the caller remains owned by the caller.
