---
title: "Get started with graft"
description: >
  Turn three ordinary tables into reviewed, connected, versioned knowledge in
  an empty local store.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Most projects do not begin with a knowledge store. They begin with tables: a
directory of people, a list of organizations, and a table that connects them.
This guide starts there. It creates an empty in-memory store, rejects a broken
relationship before writing anything, accepts corrected records, preserves a
later update as a second revision, and then hands that accepted knowledge to an
agent as read-only tools.

A pinned view lets a later task read the same accepted versions even after a
correction. Validation checks the declared contract; it does not establish
factual truth or application permission. For text-rich records, continue with
[narrative reuse](ecosystem.html) after this table example.

Install with `pak::pak("JamesHWade/graft")` before running the guide. All evaluated
chunks below use local synthetic records. The optional live-chat call is shown
but not evaluated and requires model credentials. The store closes at the end.

## Load a table contract

The package includes a small team-directory contract written with
[data-dict](https://data-dict.tidyverse.org/). This guide uses its resolved JSON
form, so the runnable lesson needs only R and graft. In a real authoring
workflow, write the YAML, validate it with the data-dict CLI, and run
`data-dict export-spec` once. The human-authored YAML is included beside the
resolved file.

```{r contract}
library(graft)

dictionary_path <- system.file(
  "extdata",
  "team-directory.data-dict.json",
  package = "graft",
  mustWork = TRUE
)

schema <- graft_schema(dictionary_path)

schema@name
names(schema@classes)
```

The three class names are the three table names: `organization`, `person`, and
`employment`. The contract defines required columns, stable string IDs, and
the two foreign keys in `employment`. IDs are strings chosen by the caller and
must be globally unique within the store; the `person:` and `org:` prefixes are
readable conventions used by this example. The `@` operator reads public
properties from graft's immutable S7 contract and plan objects.

## Create an empty store

`graft_open()` creates and initializes the database when it does not already
exist. Here the database lives only in memory and starts with no accepted
records.

```{r open, message = FALSE}
store <- graft_open(
  schema,
  path = ":memory:",
  okf = "disabled"
)
```

Use a DuckDB filename instead of `":memory:"` when the knowledge should persist
between R sessions. `okf = "disabled"` leaves the optional readable file
projection off so this first lesson stays focused on the store and its history.

## Catch a missing relationship before writing

Suppose an import contains Lois Lane and an employment row, but the referenced
organization is absent.

```{r invalid-candidate}
invalid_records <- list(
  person = data.frame(
    id = "person:lois-lane",
    full_name = "Lois Lane",
    job_title = "Reporter"
  ),
  employment = data.frame(
    id = "employment:lois-lane:daily-planet",
    person_id = "person:lois-lane",
    organization_id = "org:missing"
  )
)

invalid_plan <- graft_plan(
  store,
  invalid_records,
  graft_provenance(
    producer = "directory-import",
    idempotency_key = "directory-invalid"
  )
)

invalid_plan@valid
invalid_plan@issues[
  , c("class", "field", "rule", "message"),
  drop = FALSE
]
```

Planning is read-only. The missing organization appears as a review issue, and
neither the person nor the employment row has been accepted. `producer` names
the source workflow. An `idempotency_key` names one retryable source event so
repeating the same accepted event does not create another revision.

## Correct and review the candidate set

Add the organization and plan all three tables together. These are ordinary
data frames; the list names tell graft which contract table each frame uses.

```{r valid-candidate}
organizations <- data.frame(
  id = "org:daily-planet",
  name = "Daily Planet"
)

people <- data.frame(
  id = "person:lois-lane",
  full_name = "Lois Lane",
  job_title = "Reporter"
)

employment <- data.frame(
  id = "employment:lois-lane:daily-planet",
  person_id = "person:lois-lane",
  organization_id = "org:daily-planet"
)

initial_records <- list(
  organization = organizations,
  person = people,
  employment = employment
)

initial_plan <- graft_plan(
  store,
  initial_records,
  graft_provenance(
    producer = "directory-import",
    run_id = "import-2026-08-01",
    idempotency_key = "directory-2026-08-01"
  )
)

initial_plan@changes[
  , c("class", "record_id", "action"),
  drop = FALSE
]
initial_plan@issues
```

The organization can satisfy the employment foreign key in the same candidate
set. `@valid` means the plan can be committed; `@changes` shows whether it
should be.

## Commit the reviewed plan

```{r first-commit}
stopifnot(initial_plan@valid)

initial_result <- graft_commit(store, initial_plan)
initial_result$inserted
```

Immediately before writing, graft checks that the plan is intact and still
matches the store and active contract. All three records and their provenance
commit in one transaction.

## Accept a later update

The next directory import changes only Lois's job title. Planning compares the
candidate with the accepted head and reports one update.

```{r update}
update_plan <- graft_plan(
  store,
  list(
    person = data.frame(
      id = "person:lois-lane",
      full_name = "Lois Lane",
      job_title = "Investigative editor"
    )
  ),
  graft_provenance(
    producer = "hr-review",
    run_id = "review-2026-08-08",
    idempotency_key = "hr-review-2026-08-08"
  )
)

update_plan@changes[
  , c("class", "record_id", "action", "changed_fields"),
  drop = FALSE
]

stopifnot(update_plan@valid)
update_result <- graft_commit(store, update_plan)
update_result$updated
```

## Read current knowledge and history

`graft_get()` returns the current accepted record. The employment record keeps
the stable IDs that connect the person to the organization.

```{r current}
current_person <- graft_get(
  store,
  "person:lois-lane",
  include = character()
)
current_person$record

current_employment <- graft_get(
  store,
  "employment:lois-lane:daily-planet",
  include = character()
)$record

current_organization <- graft_get(
  store,
  current_employment$organization_id,
  include = character()
)$record
current_organization
```

`graft_history()` retains both accepted versions. It returns newest first.

```{r history}
history <- graft_history(store, "person:lois-lane", limit = 10L)

data.frame(
  revision = history$revision_number,
  committed_at = history$committed_at,
  producer = history$producer,
  event = history$source_run_id,
  contract = substr(history$schema_build_digest, 1L, 19L),
  job_title = vapply(
    history$record,
    \(record) record$job_title,
    character(1)
  )
)
```

The current record is convenient for applications. The revision history is
the audit trail: it retains what was accepted, when, under which contract, and
from which producer event. The displayed contract value is an abbreviated
build digest; the history result retains the complete digest.

## Hand the same knowledge to an agent

Everything above was written for a person at the console. The same accepted
knowledge is also what you want an agent to read, and graft exposes it without
handing over a database connection.

First pin the boundary. `graft_snapshot()` records the accepted state as a
serializable value that holds no connection and no filesystem path, and
`graft_at()` binds it to a read-only view:

```{r snapshot}
snapshot <- graft_snapshot(store)
view <- graft_at(store, snapshot)
```

`graft_tools()` then turns that view into four [ellmer](https://ellmer.tidyverse.org/)
tool definitions backed by graft's public read functions:

```{r tools, eval = rlang::is_installed("ellmer")}
tools <- graft_tools(view)
names(tools)
```

The tools are `graft_find()`, `graft_get()`, `graft_query()`, and
`graft_history()` — search, retrieval, bounded advanced operations, and
revision history. They accept no SQL, no connection, no path, and no write
argument, so a model can read accepted knowledge and cannot alter it or reach
past it.

Each call returns its result together with the limit it applied, whether the
result was truncated, and the digest of the contract it came from:

```{r tool-call, eval = rlang::is_installed("ellmer")}
found <- tools$graft_find(query = "Lois", class = "person", limit = 5)

found$truncated
found$limit
found$result[, c("id", "class", "label")]
```

A model that received only part of an answer is told so, instead of reasoning
over a silent prefix. Because these tools are bound to `view`, later commits do
not move the ground underneath a running session; a second run against the same
snapshot reads the same knowledge.

Registering them with a chat is one line:

```r
chat <- ellmer::chat_anthropic()
chat$set_tools(tools)

chat$chat("Who works at the Daily Planet, and has that person's title changed?")
```

The agent answers from accepted records and can cite the revision history
behind them. When an agent should also propose records — and be recorded as the
producer that did — read [Work with agents](agents.html).

## Close the store

```{r close}
graft_close(store)
```

## Add richer graph meaning with LinkML

The data-dict contract was enough to validate ordinary tables and their scalar
foreign keys. Those foreign keys remain validated references; graft does not
turn them into graph traversal edges.

LinkML is the next step when a relationship needs ontology identifiers,
inheritance, polymorphic targets, or explicit semantic statement and evidence
shapes. The bundled materials contract declares `Measurement` as a semantic
statement, so accepted entity-valued measurements can become typed traversal
edges. This is deliberately a second domain: the LinkML guide first explains
how the same team-directory concepts map to LinkML, then uses materials to run
a semantic edge that the package already ships.

```{r linkml-next}
linkml_schema <- graft_schema(system.file(
  "extdata",
  "materials.graft.json",
  package = "graft",
  mustWork = TRUE
))

linkml_schema@classes$Measurement@role
linkml_schema@classes$Measurement@statement_shape
```

Read [the data-dict guide](data-dict-schema.html) to adapt an existing table
contract. Next, deepen the package workflow with [change
control](knowledge-change-control.html) and [retrieval and
history](retrieval.html). Read [Work with agents](agents.html) to give a model
bounded reads and a reviewed proposal path. Then read the [LinkML
guide](linkml-schema.html) when the domain needs richer semantic structure.
