---
title: "Getting started with graft"
description: >
  Understand why graft exists and follow a complete materials example from a
  compiled LinkML contract to a source-backed claim and bounded retrieval.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

graft is for R workflows in which knowledge must remain inspectable after it
has been collected, reconciled, and reused. It keeps domain records, claims,
sources, and evidence in ordinary relational tables while enforcing a semantic
contract around them.

This matters when a workflow needs to answer more than "which rows match?" For
example:

- Is this material the same one observed in an earlier run?
- Is a statement a source-faithful narrative claim or a normalized semantic
  assertion?
- Which stored source and exact locator support or challenge the claim?
- Can an analyst or language model retrieve the answer without accessing
  arbitrary SQL or silently collecting an unbounded result?

graft addresses these questions without requiring a separate graph database.
It compiles a [LinkML](https://linkml.io/) domain schema into a portable
manifest, uses that manifest to create and protect a DuckDB store, and exposes
the stored knowledge through R, dbplyr, bounded graph projections, and optional
ellmer tools.

## The mental model

The main workflow is:

> LinkML schema -> compiled `.graft.json` manifest -> DuckDB store -> bounded R
> and agent-facing retrieval

Each part has one job:

1. **The LinkML schema describes your domain.** You define concrete classes
   such as materials, sources, claims, and evidence by extending graft's core
   record roles.
2. **The compiled manifest is the runtime contract.** It contains the resolved
   classes, fields, relational mapping, identity rules, validation invariants,
   graph projections, and schema fingerprints. Commit it with your project.
3. **DuckDB stores the records.** The database remains table-native, portable,
   and available to familiar DBI and dbplyr workflows.
4. **graft controls knowledge access.** Retrieval is limited to
   manifest-declared classes and fields. Collected results are explicitly
   bounded and carry the active schema digest.

Python and `linkml-runtime` are used only for step 2. Loading an existing
manifest and performing normal storage or retrieval both run entirely in R.

## Installation

graft is currently available from GitHub:

```{r, eval = FALSE}
pak::pak("JamesHWade/graft")
```

## Start from a compiled contract

```{r setup}
library(graft)
```

The package includes a small materials contract for this guide. Loading the
compiled manifest does not initialize Python:

```{r load-schema}
manifest <- system.file(
  "extdata",
  "materials.graft.json",
  package = "graft"
)
schema <- kg_schema(manifest)
schema
```

Inspect the contract before creating a store:

```{r inspect-schema}
kg_classes(schema)
kg_slots(schema, "Claim")
```

The `role` and `statement_shape` columns are important. They let graft treat
domain-specific classes consistently without pretending that every record is
the same kind of thing:

- nodes are the subjects and objects a workflow reasons about;
- narrative statements preserve source-faithful language;
- semantic statements express an intentional subject-predicate-object
  assertion;
- sources describe where information came from;
- evidence connects a statement to a stored source and locator.

## Create a store

For exploration, use an in-memory DuckDB database. For a durable store, replace
`":memory:"` with a file path.

```{r create-store}
store <- kg_connect_duckdb(schema, ":memory:")
kg_init(store)
store
```

`kg_init()` creates the tables and graph views declared by the manifest, plus
graft's private metadata tables. Calling it again is safe. If an existing store
was initialized with a structurally different manifest, graft reports a schema
mismatch rather than silently changing the meaning of its data.

## Ingest records with provenance

Ingestion happens in atomic batches. A batch identifies the producer and
provides an optional idempotency key, so rerunning the same stage does not
duplicate its observations.

First ingest one material and one source:

```{r ingest-foundations}
foundations <- list(
  Material = data.frame(
    preferred_name = "Linear low-density polyethylene (LLDPE)",
    description = "A polyethylene with controlled short-chain branching.",
    cas_number = "9002-88-4"
  ),
  Source = data.frame(
    uri = "https://example.org/lldpe-study",
    title = "Controlled DSC study of LLDPE crystallinity",
    doi = "10.1000/lldpe.dsc"
  )
)

kg_ingest(
  store,
  kg_batch(
    producer = "getting-started",
    source_run_id = "materials-demo",
    idempotency_key = "foundations-v1"
  ),
  foundations
)
```

graft minted internal IDs and registered the schema-declared external
identifiers. Exact lookup resolves normalized external values back to the
stable records:

```{r resolve-identities}
material_id <- kg_lookup(
  store,
  namespace = "cas",
  value = "CAS: 9002-88-4"
)$record_id[[1]]

source_id <- kg_lookup(
  store,
  namespace = "doi",
  value = "https://doi.org/10.1000/LLDPE.DSC"
)$record_id[[1]]
```

Now add a claim about the material. A multivalued LinkML slot, such as `about`,
is represented as a list-column:

```{r ingest-claim}
kg_write(
  store,
  kg_batch(
    producer = "getting-started",
    source_run_id = "materials-demo",
    idempotency_key = "claim-v1"
  ),
  class = "Claim",
  records = data.frame(
    statement_text = paste(
      "A controlled DSC experiment measured 37% crystallinity",
      "for the LLDPE sample."
    ),
    primary_subject = material_id,
    claim_type = "finding",
    polarity = "positive",
    confidence = 0.95,
    status = "active",
    about = I(list(material_id))
  )
)

claim_id <- kg_find(
  store,
  "crystallinity",
  class = "Claim",
  limit = 1
)$id[[1]]
```

Finally, connect the claim to the exact source passage:

```{r ingest-evidence}
kg_write(
  store,
  kg_batch(
    producer = "getting-started",
    source_run_id = "materials-demo",
    idempotency_key = "evidence-v1"
  ),
  class = "Evidence",
  records = data.frame(
    statement_id = claim_id,
    source_id = source_id,
    support_type = "supports",
    locator_type = "other",
    locator_value = "table 1",
    page_start = 5L,
    page_end = 5L,
    excerpt = "Crystallinity (%) for the LLDPE sample: 37."
  )
)
```

These writes are separate because the example lets graft mint each ID before a
later record refers to it. Producers that already have stable graft IDs can
submit all related classes in one atomic `kg_ingest()` call.

## Retrieve an answer and its evidence

Search is useful for discovery:

```{r search}
kg_find(store, "LLDPE", limit = 5)
```

Hydration starts from one stable ID and returns bounded related records:

```{r hydrate}
material <- kg_get(store, material_id)
material

material$record[c("preferred_name", "cas_number")]
```

Claims remain assertions rather than being collapsed into a single truth.
Their status, polarity, qualifiers, and evidence stay available for inspection:

```{r claims}
claims <- kg_claims(store, material_id)
claims[c("statement_text", "confidence", "status")]

evidence <- kg_evidence(store, claim_id)
evidence[c(
  "support_type",
  "source_title",
  "locator_value",
  "excerpt"
)]
```

This distinction is deliberate. A narrative claim records what a source says.
A semantic statement records an intentional normalized assertion. Evidence can
support or contradict either statement shape, and graft can return competing
claims without adjudicating them on the user's behalf.

## Choose the right retrieval surface

graft offers different interfaces for different jobs:

| Need | Interface | Collection behavior |
|---|---|---|
| Work with one class using dplyr | `kg_records()` | Lazy |
| Discover records by declared text fields | `kg_find()` | Bounded |
| Resolve an external identifier exactly | `kg_lookup()` | Bounded |
| Hydrate one record and related knowledge | `kg_get()` | Bounded |
| Inspect claims and stored citations | `kg_claims()`, `kg_evidence()` | Bounded |
| Run a validated structured query | `kg_select()` | Bounded |
| Explore a graph neighborhood | `kg_neighbors()` | Bounded |

For example, `kg_records()` returns a lazy dbplyr table:

```{r lazy-table}
kg_records(store, "Claim") |>
  dplyr::select(id, statement_text, confidence) |>
  dplyr::collect()
```

`kg_select()` is the safer collected interface when a caller should choose
fields and filters but must not provide SQL:

```{r structured-query}
kg_select(
  store,
  class = "Claim",
  fields = c("id", "statement_text", "confidence"),
  filters = list(list(
    field = "confidence",
    operator = "gte",
    value = 0.9
  )),
  limit = 10
)
```

## Use graft with a language model

If the `ellmer` package is installed, `kg_tools()` creates six read-only tools
that capture one initialized store:

```{r, eval = FALSE}
chat <- ellmer::chat_anthropic()
chat$set_tools(kg_tools(store))
```

The tools expose the same manifest-controlled, bounded retrieval surfaces used
above. They do not accept SQL, file paths, URLs, or network options. Their
results include truncation metadata and the store's schema digest, so a caller
can tell whether an answer was based on a complete bounded result and which
contract shaped it.

graft does not make model output trustworthy by itself. It supplies the
mechanics for retrieving stored claims and citation material without inventing
sources or bypassing the domain contract.

## Bring your own domain

The installed example's LinkML source is available as a starting point:

```{r example-schema-path}
system.file(
  "extdata",
  "materials.linkml.yaml",
  package = "graft"
)
```

Your concrete classes extend graft's core roles:

- `GraftNode`
- `GraftEdge`
- `GraftNarrativeStatement`
- `GraftSemanticStatement`
- `GraftSource`
- `GraftEvidence`
- `GraftMention`
- `GraftMetadata`

Use annotations in your LinkML schema to declare identity policies, external
identifier namespaces, search fields, label fields, origin keys, and semantic
qualifiers. Compile the schema once:

```{r, eval = FALSE}
kg_compile_schema(
  "my-domain.linkml.yaml",
  "my-domain.graft.json"
)
```

Commit both the source schema and compiled manifest. Application and production
environments can then load the manifest without Python:

```{r, eval = FALSE}
schema <- kg_schema("my-domain.graft.json")
store <- kg_connect_duckdb(schema, "knowledge.duckdb")
kg_init(store)
```

## When graft is a good fit

Use graft when domain knowledge accumulates across runs and must preserve
identity, claims, evidence, and a versioned semantic contract while remaining
comfortable to query from R.

graft is not a replacement for LinkML schema design, a general-purpose graph
database, vector similarity search, or source collection. It sits between
those concerns: producers supply domain records and evidence, LinkML defines
their meaning, DuckDB stores them as tables, and graft keeps identity,
provenance, graph projection, and retrieval behavior aligned with the compiled
contract.

```{r cleanup, include = FALSE}
kg_disconnect(store)
```
