---
title: "Get started with graft"
description: >
  Accept one reviewable knowledge change, then retrieve its current record and
  immutable history in about ten minutes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide follows the shortest complete graft workflow:

```text
contract -> provenance -> read-only plan -> review -> atomic commit -> retrieval
```

It uses a compiled contract included with the package, a temporary local
store, and no model provider. At the end you will have accepted connected
records, inspected the exact proposed change, and recovered its history.

## 1. Load the contract and open a store

```{r open}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "personinfo.graft.json",
  package = "graft"
))

store <- graft_open(
  schema,
  path = tempfile(fileext = ".duckdb"),
  okf = "disabled"
)
```

The bundled contract was compiled from LinkML and defines the classes, required
fields, stable identifiers, and relationships. Loading a compiled
`.graft.json` contract and operating a store are R-only. Source contracts may
come from LinkML or the strict data-dict table profile.

## 2. Describe the candidate change

Candidate records are a named list of data frames. The names are concrete
classes in the contract. List-columns represent multivalued fields.

```{r candidate}
records <- list(
  Organization = data.frame(
    id = "org:daily-planet",
    name = "Daily Planet"
  ),
  Person = data.frame(
    id = "person:clark-kent",
    full_name = "Clark Kent",
    employed_by = I(list("org:daily-planet"))
  )
)

origin <- graft_provenance(
  producer = "directory-import",
  run_id = "run-42",
  idempotency_key = "daily-planet-v1",
  metadata = list(source = "staff-directory.csv")
)
```

Provenance names the producer event. The idempotency key makes an intentional
retry return the original accepted result instead of creating another
observation.

## 3. Plan and review without writing

```{r plan}
plan <- graft_plan(store, records, origin)

plan@valid
plan@changes
plan@issues
```

Planning normalizes values, resolves identity, validates the relationship,
and compares the candidates with current accepted heads. It writes no accepted
records or provenance.

Review `plan@changes` even when `plan@valid` is `TRUE`. Valid means the plan
can be committed; the changes show whether it should be. Issues retain source
class and row coordinates so invalid inputs can be corrected together.

## 4. Accept the complete plan atomically

```{r commit}
if (plan@valid) {
  result <- graft_commit(store, plan)
}

result
```

Immediately before writing, graft verifies that the plan is intact and still
matches the store, active contract, and expected record heads. The records,
provenance, and identity decisions then commit together. A stale, altered, or
invalid plan fails without a partial acceptance.

When the producer is already authorized to accept its own results,
`graft_ingest(store, records, origin)` is shorthand for planning and
immediately committing a valid plan. It is not a second write path.

## 5. Retrieve the accepted view and its history

```{r retrieve}
person <- graft_get(store, "person:clark-kent")
person$record

graft_find(store, "Clark", class = "Person", limit = 10)

history <- graft_history(store, "person:clark-kent", limit = 20)
history[, c("batch_id", "changed_fields", "record")]
```

`graft_get()` reads one current accepted record. `graft_find()` searches public
fields declared by the contract. `graft_history()` reads immutable revisions
in deterministic newest-first order. Current state and history are two views
of the same accepted ledger.

Advanced retrieval stays behind fixed operations. For example, inspect the
connected organization without exposing arbitrary SQL:

```{r query}
graft_query(
  store,
  operation = "neighbors",
  request = list(
    id = "person:clark-kent",
    projection = "semantic",
    hops = 1,
    max_nodes = 25,
    max_edges = 50
  )
)
```

## 6. Close the store

```{r close}
graft_close(store)
```

Closing is idempotent. graft closes connections it created; a connection
supplied by the caller remains owned by the caller.

## Where to go next

- [Architecture](architecture.html) explains the authoritative ledger,
  derived projections, and selective use of S7.
- [LinkML contract](linkml-schema.html) covers compilation and schema
  inspection.
- [data-dict contract](data-dict-schema.html) covers the optional table-first
  adapter, resolved JSON, and its explicit semantic limits.
- [Change control](knowledge-change-control.html) covers stale plans, updates,
  history, and integrity checks.
- [Retrieval and history](retrieval.html) maps each read function to its job.
- [Open knowledge](open-knowledge-format.html) adds a readable working tree
  whose edits return through the same plan and commit boundary.
