---
title: "Get started with graft"
description: >
  Turn three ordinary tables into reviewed, connected, versioned knowledge in
  an empty local store.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Most projects do not begin with a knowledge store. They begin with tables: a
directory of people, a list of organizations, and a table that connects them.
This guide starts there. It creates an empty in-memory store, rejects a broken
relationship before writing anything, accepts corrected records, and preserves
a later update as a second revision.

## Load a table contract

The package includes a small team-directory contract written with
[data-dict](https://data-dict.tidyverse.org/). This guide uses its resolved JSON
form, so the runnable lesson needs only R and graft. In a real authoring
workflow, write the YAML, validate it with the data-dict CLI, and run
`data-dict export-spec` once. The human-authored YAML is included beside the
resolved file.

```{r contract}
library(graft)

dictionary_path <- system.file(
  "extdata",
  "team-directory.data-dict.json",
  package = "graft",
  mustWork = TRUE
)

schema <- graft_schema(dictionary_path)

schema@name
names(schema@classes)
```

The three class names are the three table names: `organization`, `person`, and
`employment`. The contract defines required columns, stable string IDs, and
the two foreign keys in `employment`. IDs are strings chosen by the caller and
must be globally unique within the store; the `person:` and `org:` prefixes are
readable conventions used by this example. The `@` operator reads public
properties from graft's immutable S7 contract and plan objects.

## Create an empty store

`graft_open()` creates and initializes the database when it does not already
exist. Here the database lives only in memory and starts with no accepted
records.

```{r open, message = FALSE}
store <- graft_open(
  schema,
  path = ":memory:",
  okf = "disabled"
)
```

Use a DuckDB filename instead of `":memory:"` when the knowledge should persist
between R sessions. `okf = "disabled"` leaves the optional readable file
projection off so this first lesson stays focused on the store and its history.

## Catch a missing relationship before writing

Suppose an import contains Lois Lane and an employment row, but the referenced
organization is absent.

```{r invalid-candidate}
invalid_records <- list(
  person = data.frame(
    id = "person:lois-lane",
    full_name = "Lois Lane",
    job_title = "Reporter"
  ),
  employment = data.frame(
    id = "employment:lois-lane:daily-planet",
    person_id = "person:lois-lane",
    organization_id = "org:missing"
  )
)

invalid_plan <- graft_plan(
  store,
  invalid_records,
  graft_provenance(
    producer = "directory-import",
    idempotency_key = "directory-invalid"
  )
)

invalid_plan@valid
invalid_plan@issues[
  , c("class", "field", "rule", "message"),
  drop = FALSE
]
```

Planning is read-only. The missing organization appears as a review issue, and
neither the person nor the employment row has been accepted. `producer` names
the source workflow. An `idempotency_key` names one retryable source event so
repeating the same accepted event does not create another revision.

## Correct and review the candidate set

Add the organization and plan all three tables together. These are ordinary
data frames; the list names tell graft which contract table each frame uses.

```{r valid-candidate}
organizations <- data.frame(
  id = "org:daily-planet",
  name = "Daily Planet"
)

people <- data.frame(
  id = "person:lois-lane",
  full_name = "Lois Lane",
  job_title = "Reporter"
)

employment <- data.frame(
  id = "employment:lois-lane:daily-planet",
  person_id = "person:lois-lane",
  organization_id = "org:daily-planet"
)

initial_records <- list(
  organization = organizations,
  person = people,
  employment = employment
)

initial_plan <- graft_plan(
  store,
  initial_records,
  graft_provenance(
    producer = "directory-import",
    run_id = "import-2026-08-01",
    idempotency_key = "directory-2026-08-01"
  )
)

initial_plan@changes[
  , c("class", "record_id", "action"),
  drop = FALSE
]
initial_plan@issues
```

The organization can satisfy the employment foreign key in the same candidate
set. `@valid` means the plan can be committed; `@changes` shows whether it
should be.

## Commit the reviewed plan

```{r first-commit}
stopifnot(initial_plan@valid)

initial_result <- graft_commit(store, initial_plan)
initial_result$inserted
```

Immediately before writing, graft checks that the plan is intact and still
matches the store and active contract. All three records and their provenance
commit in one transaction.

## Accept a later update

The next directory import changes only Lois's job title. Planning compares the
candidate with the accepted head and reports one update.

```{r update}
update_plan <- graft_plan(
  store,
  list(
    person = data.frame(
      id = "person:lois-lane",
      full_name = "Lois Lane",
      job_title = "Investigative editor"
    )
  ),
  graft_provenance(
    producer = "hr-review",
    run_id = "review-2026-08-08",
    idempotency_key = "hr-review-2026-08-08"
  )
)

update_plan@changes[
  , c("class", "record_id", "action", "changed_fields"),
  drop = FALSE
]

stopifnot(update_plan@valid)
update_result <- graft_commit(store, update_plan)
update_result$updated
```

## Read current knowledge and history

`graft_get()` returns the current accepted record. The employment record keeps
the stable IDs that connect the person to the organization.

```{r current}
current_person <- graft_get(
  store,
  "person:lois-lane",
  include = character()
)
current_person$record

current_employment <- graft_get(
  store,
  "employment:lois-lane:daily-planet",
  include = character()
)$record

current_organization <- graft_get(
  store,
  current_employment$organization_id,
  include = character()
)$record
current_organization
```

`graft_history()` retains both accepted versions. It returns newest first.

```{r history}
history <- graft_history(store, "person:lois-lane", limit = 10L)

data.frame(
  revision = history$revision_number,
  committed_at = history$committed_at,
  producer = history$producer,
  event = history$source_run_id,
  contract = substr(history$schema_build_digest, 1L, 19L),
  job_title = vapply(
    history$record,
    \(record) record$job_title,
    character(1)
  )
)
```

The current record is convenient for applications. The revision history is
the audit trail: it retains what was accepted, when, under which contract, and
from which producer event. The displayed contract value is an abbreviated
build digest; the history result retains the complete digest.

## Close the store

```{r close}
graft_close(store)
```

## Add richer graph meaning with LinkML

The data-dict contract was enough to validate ordinary tables and their scalar
foreign keys. Those foreign keys remain validated references; graft does not
turn them into graph traversal edges.

LinkML is the next step when a relationship needs ontology identifiers,
inheritance, polymorphic targets, or explicit semantic statement and evidence
shapes. The bundled materials contract declares `Measurement` as a semantic
statement, so accepted entity-valued measurements can become typed traversal
edges. This is deliberately a second domain: the LinkML guide first explains
how the same team-directory concepts map to LinkML, then uses materials to run
a semantic edge that the package already ships.

```{r linkml-next}
linkml_schema <- graft_schema(system.file(
  "extdata",
  "materials.graft.json",
  package = "graft",
  mustWork = TRUE
))

linkml_schema@classes$Measurement@role
linkml_schema@classes$Measurement@statement_shape
```

Read [the data-dict guide](data-dict-schema.html) to adapt an existing table
contract. Next, deepen the package workflow with [change
control](knowledge-change-control.html) and [retrieval and
history](retrieval.html). Then read the [LinkML guide](linkml-schema.html) when
the domain needs richer semantic structure.
