---
title: "Govern knowledge changes"
description: >
  Inspect what changed, recover accepted state at a commit boundary, and apply
  a reviewed additive schema migration.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Govern knowledge changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A current-state table can answer "what do we believe now?" It cannot, on its
own, answer the questions that make the result dependable:

- What changed, and which workflow run changed it?
- What had the store accepted before that change?
- Which exact schema governed each revision?
- Can the contract evolve without silently changing stored knowledge?

graft keeps the typed tables used for ordinary queries and an append-only
revision ledger used for these questions. Ingestion batches connect each
accepted change to its producer, schema, and commit order. Schema evolution is
a separate plan-review-apply operation.

This guide runs the complete loop with two small contracts included in the
package. Both are already compiled, so the walkthrough does not require
Python.

## Create a version 1 store

```{r setup}
library(graft)

manifest_v1 <- system.file(
  "extdata",
  "change-control-v1.graft.json",
  package = "graft"
)
schema_v1 <- kg_schema(manifest_v1)
store <- kg_connect_duckdb(schema_v1, ":memory:")
kg_init(store)
```

The first batch records an initiative at the point it enters concept review.
The producer-side run identifier and idempotency key make that provenance and
replay boundary explicit.

```{r baseline}
baseline <- kg_write(
  store,
  kg_batch(
    producer = "portfolio-review",
    source_run_id = "daily-brief-001",
    idempotency_key = "firefly-v1"
  ),
  class = "Initiative",
  records = data.frame(
    preferred_name = "Project Firefly",
    status = "concept review",
    portfolio_id = "PF-001"
  )
)

initiative_id <- kg_lookup(
  store,
  namespace = "portfolio",
  value = "PF-001"
)$record_id[[1]]
```

A later batch accepts a new status for the same schema-declared identity:

```{r update}
approved <- kg_write(
  store,
  kg_batch(
    producer = "portfolio-review",
    source_run_id = "daily-brief-002",
    idempotency_key = "firefly-v2"
  ),
  class = "Initiative",
  records = data.frame(
    preferred_name = "Project Firefly",
    status = "prototype approved",
    portfolio_id = "PF-001"
  )
)
```

The typed table remains the convenient current-state projection:

```{r current-state}
kg_records(store, "Initiative") |>
  dplyr::select(preferred_name, status, portfolio_id) |>
  dplyr::collect()
```

## Inspect batches and changes

`kg_batches()` returns committed workflow batches in newest-first commit order.
Its filters and result limits make it suitable for routine status views rather
than an unbounded audit-table dump.

```{r batches}
kg_batches(store)[
  c("commit_order", "producer", "source_run_id", "status")
]
```

`kg_changes()` returns accepted inserts and updates. Each row identifies the
batch and schema build that governed the revision. Parsed records and changed
field names are exposed as list-columns; raw ledger payloads and content
digests are not.

```{r changes}
kg_changes(store, record_id = initiative_id)[
  c(
    "commit_order",
    "revision_number",
    "operation",
    "changed_fields",
    "batch_id",
    "schema_build_digest"
  )
]
```

`kg_history()` narrows the same ledger to one stable record:

```{r history}
kg_history(store, initiative_id)[
  c("revision_number", "operation", "changed_fields", "recorded_at")
]
```

## Recover state at a commit boundary

Pass a committed batch ID to recover what Graft had accepted when that batch
completed. A scalar `POSIXt` value is also supported; Graft first resolves it
to the latest committed batch order, so equal or imprecise timestamps do not
reorder transactions.

```{r as-of}
at_baseline <- kg_history(
  store,
  initiative_id,
  as_of = baseline$batch_id,
  limit = 1
)

at_baseline$record[[1]][
  c("preferred_name", "status", "portfolio_id")
]
```

This is **system time**: when Graft committed a version. It is different from
**domain-valid time**: when a claim is true in the world. Statement schemas can
model domain validity with `valid_from` and `valid_to`; those remain
user-supplied facts. Revision history should not be used as a substitute for
them.

## Plan and review an additive migration

Version 2 of the example contract adds an optional `decision_note` field.
Loading it does not change the store.

```{r load-v2}
manifest_v2 <- system.file(
  "extdata",
  "change-control-v2.graft.json",
  package = "graft"
)
schema_v2 <- kg_schema(manifest_v2)

kg_schema_diff(schema_v1, schema_v2)
```

The `additive` classification means this difference is supported by the current
migration path.

Planning is also read-only. The result is deterministic and bound to this
store's identity, format version, and active schema build. Review the
classification, every classified change, and the declarative operations before
approval.

```{r plan-migration}
plan <- kg_plan_migration(store, schema_v2)
plan

plan$changes[
  c("path", "object_type", "classification", "rule")
]

lapply(plan$operations, function(operation) {
  list(
    kind = operation$kind,
    table = operation$table,
    column = operation$column$name
  )
})
```

The current migration engine accepts only compatible and supported additive
changes. Required fields, removals, type changes, normalization changes,
custom transformations, and other risky changes are reported but refused.
There is no force flag.

Once the plan has been reviewed, apply that exact object:

```{r apply-migration}
kg_apply_migration(store, plan)

kg_store_info(store)[
  c(
    "store_format_version",
    "required_store_format_version",
    "active_build_digest",
    "history_complete"
  )
]
```

Application revalidates the plan and its store preconditions, performs the DDL,
registers and activates the exact target manifest, rebuilds generated graph
views, and records the migration in one transaction.

## Continue writing under the new contract

The same stable initiative can now use the added field. Unchanged fields do not
create false differences merely because the governing schema changed.

```{r post-migration-write}
with_decision <- kg_write(
  store,
  kg_batch(
    producer = "portfolio-review",
    source_run_id = "daily-brief-003",
    idempotency_key = "firefly-v3"
  ),
  class = "Initiative",
  records = data.frame(
    preferred_name = "Project Firefly",
    status = "prototype approved",
    portfolio_id = "PF-001",
    decision_note = "Advance with a reversible prototype."
  )
)

kg_records(store, "Initiative") |>
  dplyr::select(preferred_name, status, decision_note) |>
  dplyr::collect()
```

History spans both schema builds and applies each historical manifest's
sensitivity rules when hydrating records:

```{r cross-schema-history}
kg_changes(store, record_id = initiative_id)[
  c(
    "revision_number",
    "changed_fields",
    "schema_build_digest"
  )
]
```

## Check integrity

`kg_check_store()` validates relationships among batches, revisions, heads,
schema versions, and current typed tables. Deep checking also re-digests every
revision payload and compares current records with their revision heads.

```{r integrity}
check <- kg_check_store(store, deep = TRUE)
check
check$issues
```

Direct writes to client tables are unsupported because they bypass the ledger.
A deep check reports that kind of current-state drift.

```{r cleanup, include = FALSE}
kg_disconnect(store)
```

## Store format 2 is a clean cutover

Knowledge history begins with store format 2. Stores created by earlier
development versions are rejected with a classed format error; Graft does not
silently upgrade them or run in a legacy mode. During this pre-release phase,
recreate an older development store from its source records under the current
contract.

This boundary keeps every format 2 store honest about the same guarantee: its
accepted current state has complete revision history from initialization
onward.
