---
title: "Add graph semantics with LinkML"
description: >
  Use LinkML when a table contract needs ontology identifiers, inheritance, or
  semantic statements that produce a bounded graph projection.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add graph semantics with LinkML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(graft)
```

Start with data-dict when the domain is naturally a set of tables. It gives
Graft table and column descriptions, primary keys, supported value rules, and
scalar foreign keys that can be checked before records are accepted.

LinkML is the next step when the contract must express meaning beyond those
tables. It can define ontology identifiers, inheritance, polymorphic ranges,
and Graft semantic-statement roles. Both providers compile to the same
`.graft.json` runtime contract and use the same plan, commit, history, and
retrieval functions.

There is an important distinction: a checked reference is not automatically a
graph edge. Graft validates ordinary LinkML object-reference slots, just as it
validates supported data-dict foreign keys. The semantic graph projection is
driven by classes declared as graph-producing edges or semantic statements.

| Need | Start with |
|---|---|
| Tables, columns, keys, enums, and checked scalar foreign keys | data-dict |
| Class and slot ontology identifiers | LinkML |
| Reusable inheritance or polymorphic references | LinkML |
| Subject-predicate-object statements in the semantic graph | LinkML with Graft semantic roles |

## Add ontology identifiers and inheritance

An ordinary LinkML schema can introduce shared fields through inheritance and
attach stable ontology identifiers to classes and slots:

```yaml
id: https://w3id.org/example/team-directory
name: team_directory
version: 0.1.0
prefixes:
  linkml: https://w3id.org/linkml/
  sdo: https://schema.org/
  team: https://w3id.org/example/team-directory/
default_prefix: team
default_range: string
imports:
  - linkml:types

classes:
  NamedThing:
    abstract: true
    attributes:
      id:
        identifier: true
        required: true
      name:
        required: true

  Person:
    is_a: NamedThing
    class_uri: sdo:Person
    attributes:
      works_for:
        range: Organization
        inlined: false

  Organization:
    is_a: NamedThing
    class_uri: sdo:Organization
```

This contract reuses identity and name fields and validates `works_for` as an
organization reference. That reference alone does not create a traversal edge.
Use a graph-producing statement when the relationship must participate in the
semantic graph.

## Declare a semantic statement

Graft ships a small LinkML core for schemas that need record and statement
roles. To run a complete graph example without introducing another
team-directory fixture, the rest of this guide switches deliberately to the
bundled materials domain. It defines nodes and a semantic statement:

```yaml
classes:
  Material:
    is_a: GraftNode
    slots:
      - preferred_name
      - description
      - cas_number

  Measurement:
    is_a: GraftSemanticStatement
    slots:
      - measurement_method
      - temperature
```

`GraftSemanticStatement` supplies `subject`, `predicate`, `object_entity`, and
`object_value`. A concrete statement must have exactly one object
representation. Accepted entity-valued statements produce the semantic edges
used by bounded neighbor retrieval. Narrative statements remain searchable
claims but are excluded from that semantic edge projection.

This richer behavior is why LinkML follows, rather than replaces, the simpler
data-dict path. Use it when the domain actually needs semantic graph meaning.

## Compile once, then run in R

`graft_schema()` compiles LinkML YAML and its import closure into a canonical
`.graft.json` manifest:

```{r compile, eval = FALSE}
library(graft)

schema <- graft_schema(
  system.file(
    "extdata",
    "materials.linkml.yaml",
    package = "graft",
    mustWork = TRUE
  ),
  output = "materials.graft.json"
)
```

Compilation requires Python and `linkml-runtime`. Graft rejects source
semantics that its runtime contract cannot preserve instead of silently
downgrading them. [Contract compiler boundaries](contract-compilers.html)
lists the supported LinkML profile and the exact compilation dependencies.

Loading a compiled manifest and operating a store are R-only:

```{r package-contract}
schema <- graft_schema(system.file(
  "extdata",
  "materials.graft.json",
  package = "graft",
  mustWork = TRUE
))
```

Commit the LinkML source and compiled manifest together when users or deployed
workflows should run without Python.

## Inspect what the contract enables

The typed schema properties expose the parts most applications need:

```{r inspect}
schema@name
schema@version
schema@structural_digest
names(schema@classes)

material <- schema@classes$Material
material@role
material@label_slot
material@search_slots

measurement <- schema@classes$Measurement
measurement@role
measurement@statement_shape
measurement@slots$subject@object_reference
measurement@slots$object_entity@object_reference
```

The complete manifest is available as `schema@manifest` for lower-level
tooling. Domain records remain ordinary data frames; the S7 schema object
protects the compiled contract and its digests.

## Accept and traverse semantic statements

The materials contract can accept two nodes and a semantic statement in one
plan:

```{r records, message = FALSE}
store <- graft_open(schema, ":memory:", okf = "disabled")

polyethylene_id <- "graft:40X0M02WR9Z55QWFDJYTNS3KX8"
water_id <- "graft:64W8ZFH68S76XQ2MWW0H46Q9Y6"

records <- list(
  Material = data.frame(
    id = c(polyethylene_id, water_id),
    preferred_name = c("Polyethylene", "Water"),
    cas_number = c("9002-88-4", "7732-18-5")
  ),
  Measurement = data.frame(
    id = "graft:0D6WMEBSAMAS7JRQWGT5W44ASR",
    subject = polyethylene_id,
    predicate = "materials:testedWith",
    object_entity = water_id,
    measurement_method = "immersion",
    temperature = 23
  )
)

plan <- graft_plan(
  store,
  records,
  graft_provenance(
    producer = "materials-example",
    idempotency_key = "materials-v1"
  )
)

plan@valid
plan@changes[, c("class", "record_id", "action"), drop = FALSE]
plan@issues

stopifnot(plan@valid)
result <- graft_commit(store, plan)
result$inserted
```

Because `Measurement` is a semantic statement, the bounded neighbor operation
can return its subject, object, and predicate:

```{r neighbors}
neighbors <- graft_query(
  store,
  operation = "neighbors",
  request = list(
    id = polyethylene_id,
    projection = "semantic",
    hops = 1,
    max_nodes = 25,
    max_edges = 50
  )
)

neighbors$nodes[, c("id", "label"), drop = FALSE]
neighbors$edges[, c("subject", "predicate", "object"), drop = FALSE]
```

The graph is a derived read view of accepted revisions. It is not a separate
place to insert or edit relationships.

## Change the contract explicitly

Every plan and accepted revision records the active schema digest. Editing the
LinkML source therefore creates a new contract version rather than silently
changing the meaning of existing knowledge.

During v0.1 development, Graft can register a compatible semantic contract and
rebuild its derived views. A change that transforms accepted payloads requires
rebuilding the development store from source records. Historical revisions
retain the contract digest that governed their acceptance.

```{r close}
graft_close(store)
```
