---
title: "Use a LinkML schema with graft"
description: >
  Compile an ordinary LinkML schema, write schema-shaped records to DuckDB,
  and query them from R.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use a LinkML schema with graft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article starts with an ordinary LinkML schema. It does not import graft's
core schema or use graft annotations.

The example is based on LinkML's
[PersonInfo tutorial](https://linkml.io/linkml/intro/tutorial01.html), which
introduces classes and attributes with a small `Person` model. This version
adds an `Organization` class and the tutorial's suggested employment
relationship.

## Write the schema

```{r setup}
library(graft)

schema_file <- system.file(
  "extdata",
  "personinfo.linkml.yaml",
  package = "graft"
)
```

The complete schema is:

```{r show-schema, echo = FALSE, results = "asis"}
cat(
  "```yaml\n",
  paste(readLines(schema_file, warn = FALSE), collapse = "\n"),
  "\n```\n",
  sep = ""
)
```

These are standard LinkML features:

- `classes` define `Person` and `Organization` records;
- `attributes` define their fields;
- `identifier: true` marks each class identifier;
- `required`, `range`, and `minimum_value` add constraints;
- `multivalued: true` represents repeated values; and
- `slot_uri` and `class_uri` connect local names to Schema.org terms.

There is no graft-specific authoring requirement in this file.

## Compile the schema

Compile the LinkML file into a graft manifest:

```{r compile, eval = FALSE}
kg_compile_schema(
  schema_file,
  "personinfo.graft.json"
)
```

Compilation is the only step that uses Python and `linkml-runtime`. Commit the
generated `.graft.json` file with the source schema. Applications can then load
the manifest without Python.

The package includes a compiled copy so the rest of this article is
executable:

```{r load-manifest}
manifest_file <- system.file(
  "extdata",
  "personinfo.graft.json",
  package = "graft"
)
schema <- kg_schema(manifest_file)

kg_classes(schema)
kg_slots(schema, "Person")
```

For a plain LinkML class, graft supplies conservative storage defaults:

- a concrete class becomes a node table;
- a scalar string-like `id` remains the record identifier;
- scalar attributes become columns;
- multivalued attributes become related tables;
- class-valued attributes become references;
- common text fields provide a display label and basic search fields; and
- `created_at` and `updated_at` are added to the runtime manifest.

These defaults affect the compiled manifest, not the source LinkML file.

## Create the DuckDB store

```{r create-store}
store <- suppressMessages(kg_connect_duckdb(schema, ":memory:"))
kg_init(store)
```

Use a file path instead of `":memory:"` when the store should persist across R
sessions.

## Write LinkML-shaped records

The input data use the class and attribute names from the LinkML schema. An
object reference contains the identifier of the referenced record. A
multivalued attribute uses a list-column.

```{r ingest}
records <- list(
  Organization = data.frame(
    id = "org:daily-planet",
    name = "Daily Planet"
  ),
  Person = data.frame(
    id = "person:clark-kent",
    full_name = "Clark Kent",
    aliases = I(list(c("Superman", "Kal-El"))),
    age = 35L,
    employed_by = I(list("org:daily-planet"))
  )
)

kg_ingest(
  store,
  kg_batch(
    producer = "linkml-example",
    idempotency_key = "personinfo-v1"
  ),
  records
)
```

The two classes are committed in one transaction. The reference from
`Person.employed_by` is checked against both records already in the store and
records staged in the same batch.

## Query the records

`kg_records()` returns a lazy dbplyr table:

```{r lazy-query}
people <- kg_records(store, "Person")

people |>
  dplyr::select(id, full_name, age) |>
  dplyr::collect()
```

The inferred text fields are available to `kg_find()`:

```{r search}
kg_find(store, "Clark", class = "Person", limit = 5)
```

`kg_get()` hydrates the record and its multivalued fields:

```{r hydrate}
person <- kg_get(store, "person:clark-kent")

person$record[c("id", "full_name", "age")]
person$record$aliases
person$record$employed_by
```

## Add graft semantics only when needed

Ordinary LinkML describes record structure, identifiers, ranges, cardinality,
and semantic mappings. That is enough for baseline storage and querying.

Some knowledge-store decisions are not part of a general LinkML schema. For
example, a project may want to distinguish narrative claims from normalized
semantic statements, connect claims to exact evidence locations, normalize
external identifiers, or choose weighted search fields.

For those cases, import `graft-core.linkml` and extend one of graft's optional
roles:

- `GraftNarrativeStatement`
- `GraftSemanticStatement`
- `GraftSource`
- `GraftEvidence`
- `GraftMention`
- `GraftEdge`
- `GraftMetadata`

The package's
[`materials.linkml.yaml`](https://github.com/JamesHWade/graft/blob/main/inst/extdata/materials.linkml.yaml)
shows that enriched form. Start with ordinary LinkML and add these roles only
when the application needs their behavior.

```{r cleanup, include = FALSE}
kg_disconnect(store)
```
