---
title: "Work with open knowledge"
description: >
  Synchronize a readable OKF working tree, inspect its state, and review edits
  through Graft's ordinary plan and commit contract.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work with open knowledge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The [Open Knowledge
Format](https://github.com/GoogleCloudPlatform/knowledge-catalog/tree/main/okf)
(OKF) gives accepted knowledge a human- and agent-readable working surface.
People can open Markdown, follow links, inspect provenance, and review a Git
diff without needing a database client.

Readable files are useful, but they are not a second source of truth. In graft:

- The compiled Graft manifest is the domain contract.
- The revision ledger is the authority for accepted knowledge.
- OKF is a deterministic projection and proposal surface.
- `graft_review()` and `graft_commit()` are the route from edited files back to
  accepted knowledge.

This boundary keeps filesystem edits reviewable without allowing them to
bypass identity, validation, provenance, or commit preconditions.

## Configure the working tree

A file-backed store manages a sibling OKF directory by default. For example,
`knowledge.duckdb` uses `knowledge.okf` unless `okf_path` is supplied.

```{r open}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "personinfo.graft.json",
  package = "graft"
))

store <- graft_open(
  schema,
  path = "knowledge.duckdb",
  okf = "managed"
)
```

Use `okf = "disabled"` when a store should have no managed working tree. An
explicit path is useful when the readable projection belongs inside a
repository:

```{r custom-path}
store <- graft_open(
  schema,
  path = "knowledge.duckdb",
  okf_path = "knowledge/okf"
)
```

## Synchronize accepted knowledge

After committing records, synchronize explicitly:

```{r seed}
graft_ingest(
  store,
  list(
    Organization = data.frame(
      id = "org:daily-planet",
      name = "Daily Planet"
    ),
    Person = data.frame(
      id = "person:clark-kent",
      full_name = "Clark Kent",
      employed_by = I(list("org:daily-planet"))
    )
  ),
  graft_provenance(
    producer = "directory-import",
    idempotency_key = "daily-planet-v1"
  )
)

bundle <- graft_sync(store)
bundle
```

`graft_sync()` stages and validates a complete deterministic bundle before
replacing a graft-managed destination. It never changes accepted records and
does not replace an unrelated directory.

Synchronization is separate from commit for a reason. Once a transaction has
accepted a revision, a later filesystem failure cannot make that authoritative
commit ambiguous. The working tree can simply be synchronized again.

The bundle contains a navigable index and concept documents. Public record
content is stored in readable frontmatter and prose; object references become
links; source and revision identity remain traceable. Sensitive fields are
filtered according to the contract that governs retrieval.

## Inspect state without changing it

```{r status}
status <- graft_status(store)
status$status
status$reason
```

`graft_status()` is read-only. It distinguishes these states:

- `current`: the working tree matches current accepted knowledge;
- `modified`: local edits form a proposal against the current accepted base;
- `stale`: the store has newer accepted revisions;
- `missing`: the configured working tree does not exist;
- `unconfigured`: the store has no managed OKF path; and
- `incompatible`: the bundle cannot be interpreted under the active contract.

Use `deep = TRUE` to verify the working tree's content digest. Status does not
silently repair, overwrite, or accept anything.

## Treat edits as proposals

A person or tool may edit the structured record mapping in a concept document.
The edited file is still only a proposal. Review it with explicit provenance:

```{r review}
review <- graft_review(
  store,
  provenance = graft_provenance(
    producer = "human-review",
    run_id = "review-17",
    idempotency_key = "approved-okf-edit-17"
  )
)

review@valid
review@changes
review@issues
```

`graft_review()` reads a stable snapshot, compares the proposed records with
current accepted revisions, and applies the same normalization, identity, and
contract validation as `graft_plan()`. The returned `GraftCommitPlan` is bound to
the exact bundle digest, accepted base, active schema, and store.

Only complete managed bundles can be reviewed. Removing concept documents is
not a supported way to delete knowledge. Restore the document before planning
another proposal.

## Accept through the ordinary commit boundary

```{r accept}
if (review@valid) {
  result <- graft_commit(store, review)
}
```

Committing rechecks both the ordinary plan preconditions and the OKF source
snapshot. If the files changed after review, the accepted base advanced, or
the contract changed, the commit fails without a partial write. Review the new
state and create a fresh plan.

After acceptance, synchronize the working tree to the new accepted heads:

```{r resync}
graft_sync(store)
graft_status(store)
```

The complete loop is therefore explicit:

```text
accepted revisions -> sync -> readable files -> edit -> review -> commit -> sync
```

## Give agents the accepted view

`graft_tools()` creates four read-only tool definitions backed by
`graft_find()`, `graft_get()`, `graft_query()`, and `graft_history()`:

```{r tools}
tools <- graft_tools(store)
names(tools)
```

The tool definitions expose bounded accepted retrieval, not filesystem access
or mutation. A modified OKF document remains a proposal until it passes review
and commit. The host decides which model provider, if any, receives the tools.

```{r close}
graft_close(store)
```
