---
title: "Share accepted knowledge with OKF"
description: >
  Publish current or historical Graft revisions as portable, source-linked
  Markdown for agents, Git, and documentation tools.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Share accepted knowledge with OKF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Graft and the [Open Knowledge
Format](https://github.com/GoogleCloudPlatform/knowledge-catalog/tree/main/okf)
(OKF) solve different problems. Graft validates writes against a compiled
domain contract and preserves every accepted revision. OKF packages selected
knowledge as plain Markdown and YAML that people, agents, Git, and
documentation tools can inspect.

`kg_export_okf()` connects those layers without confusing them:

| Layer | Governs | Primary representation |
|---|---|---|
| LinkML | Domain classes, slots, ranges, identity, and sensitivity | Source schema |
| Graft | Executable validation, accepted state, revisions, and retrieval | `.graft.json` manifest and DuckDB |
| OKF | Portable reading, navigation, provenance, and exchange | Markdown directory |

The LinkML-derived manifest remains authoritative for writes. An OKF bundle is
a projection of an accepted boundary, not a second writable database.

## Export current accepted state

Pass an initialized store and a destination directory:

```{r current, eval = FALSE}
library(graft)

bundle <- kg_export_okf(
  store,
  "knowledge/okf",
  limit = 5000
)

bundle
```

The result reports the directory, OKF and Graft profile versions, concept
count, exported classes, schema digest, and accepted time boundary. The
directory contains:

```text
knowledge/okf/
├── index.md
└── concepts/
    ├── index.md
    ├── Assessment/
    │   ├── index.md
    │   └── market%3Aresin-demand.md
    └── Source/
        ├── index.md
        └── source%3Aquarterly-results.md
```

Exports are deterministic for the same accepted state. Graft stages and
validates the complete bundle before installing it. It refuses to write a
partial result when `limit` is exceeded and refuses to replace an unrelated
directory. Set `overwrite = TRUE` only to replace a directory that already
identifies itself as a complete Graft OKF bundle.

Use `classes` to publish a deliberate subset:

```{r selected-classes, eval = FALSE}
kg_export_okf(
  store,
  "knowledge/market-radar",
  classes = c("Business", "Assessment", "Source")
)
```

References to records outside that selection remain visible as stable
identifiers rather than becoming broken promises about included content.

## Reproduce a historical boundary

`as_of` accepts a committed batch identifier or a scalar `POSIXt` value:

```{r historical, eval = FALSE}
quarter_close <- kg_export_okf(
  store,
  "knowledge/2026-q2-close",
  as_of = "batch-2026-q2-close"
)

morning_view <- kg_export_okf(
  store,
  "knowledge/2026-07-28-morning",
  as_of = as.POSIXct("2026-07-28 08:00:00", tz = "UTC")
)
```

Graft reconstructs each record from the last accepted revision at that
boundary. It also uses the historical manifest that governed that revision,
so fields marked sensitive at that time remain excluded from the projection.

## Preserve identity, evidence, and relationships

Every concept has standard OKF frontmatter and a namespaced `graft` extension.
The extension records:

- stable record and class identity;
- revision number and content digest;
- committed batch and source-run identity; and
- the exact schema build digest.

The readable Markdown body follows the domain contract:

- label slots become titles;
- statement or descriptive text becomes the primary narrative;
- non-sensitive scalar fields become a details table;
- object references to included records become Markdown links; and
- direct source references and statement evidence become OKF `sources` and
  footnotes.

Graft maps common lifecycle values to OKF's `draft`, `stable`, and
`deprecated` statuses. Batch producer and commit time become generation
provenance. The export does not invent human verification: approval remains
captured by the accepted Graft revision and its host workflow.

## Hand the bundle to Tempest

Tempest can read any conformant OKF bundle, including a Graft projection:

```{r tempest-handoff, eval = FALSE}
library(tempest)

knowledge <- tempest_read_okf("knowledge/okf")
tempest_okf_concepts(knowledge)

resources <- tempest_okf_resources(
  knowledge,
  include_stale = FALSE
)

evidence <- SourceStore$new()
invisible(lapply(resources, evidence$upsert_resource))

context <- tempest_okf_context(
  knowledge,
  types = c("Assessment", "Business"),
  include_stale = FALSE,
  max_concepts = 25,
  max_chars = 50000
)
```

Reading is intentionally separate from adding resources to a `SourceStore`.
This keeps the mutation visible to the host. The bounded context starts with a
trust notice and cannot grant tools, change workflow policy, approve an
artifact, or authorize an external action.

## Keep the authority boundary explicit

An OKF bundle is useful evidence, not ambient authority:

| Operation | Authority |
|---|---|
| Read Markdown and inspect metadata | OKF consumer |
| Decide which concepts enter a workflow | Host application |
| Grant model tools or authenticated connections | Tempest runtime and host policy |
| Approve a generated interpretation or action | Human or explicit host policy |
| Commit accepted knowledge | Graft write path |

This separation lets an agent carry organizational context without turning
documents into executable instructions. Even OKF Attested Computation
documents remain descriptions in Tempest; referenced code is never executed
by the reader.

Continue with [Govern knowledge changes](knowledge-change-control.html) for
revision history and schema evolution, or [Build a governed materials market
radar](market-intelligence.html) for an approval-gated application that can
publish its accepted state.
