---
title: "Persistent artifacts and selected memory"
description: >
  Save exact artifacts, record host decisions and guard current reuse.
  Understand the relationship to Commons, data-dict and the migration evidence.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent artifacts and selected memory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Your saved work should survive both the end of an R session and a change in the
software that stores it. Graft preserves reports, tables, figures and the exact
evidence selected for later workflows. Its decision journal records host reviews
and separates historical inspection from current reuse.

## Save and reopen exact content

The shared local artifact interface preserves opaque bytes and verifies them when
read. It supports trusted local files with a single writer. Application approval,
access, concurrent publication, power-loss durability and permanent erasure remain
separate requirements; saving these bytes grants no approval.

```{r}
library(graft)
path <- tempfile("artifacts-")
store <- graft_artifact_store(path, create = TRUE)
original <- graft_artifact_save(
  store, "report:daily", charToRaw("Original report"), "text/plain"
)
corrected <- graft_artifact_save(
  store, "report:daily", charToRaw("Corrected report"), "text/plain"
)
reopened <- graft_artifact_store(path)
rawToChar(graft_artifact_read(reopened, original)$bytes)
rawToChar(graft_artifact_read(reopened, corrected)$bytes)
unlink(path, recursive = TRUE)
```

References contain an identity and immutable revision. Retain them with your
workflow; the store deliberately has no automatic latest or approved selection.
This project is pre-production. Consumer contract 3 removes the graph/compiler APIs.
The experiments below remain comparative evidence, not the installed storage code.

## Pin evidence and shared meaning

A dependency is an exact artifact reference. Store dictionary and vocabulary
releases as content, then reference them alongside the evidence used in a report.
The selection preserves their bytes; it does not evaluate expressions or infer
that two measurements are comparable.

```{r}
path <- tempfile("artifact-selection-")
store <- graft_artifact_store(path, create = TRUE)
vocabulary <- graft_artifact_save(
  store, "vocabulary:lab", charToRaw('{"concept":"sample"}'), "application/json"
)
dictionary <- graft_artifact_save(
  store, "dictionary:assay", charToRaw("name: assay"), "application/yaml",
  dependencies = list(vocabulary)
)
report <- graft_artifact_save(
  store, "report:assay", charToRaw("Assay interpretation"), "text/plain",
  dependencies = list(dictionary)
)
selection <- graft_artifact_select(store, list(report))
graft_artifact_read_selection(graft_artifact_store(path), selection)$artifacts
unlink(path, recursive = TRUE)
```

The returned list includes the report, dictionary and vocabulary revisions.
`max_metadata_bytes` separately bounds encoded selection metadata (1 MiB by
default). Increase it for many long artifact identities and pass that limit on
both selection creation and reading. Similarly, open the store with a larger
`max_revision_bytes` (default 1 MiB) when saving or reading artifacts whose
dependency lists produce large revision manifests.

`max_artifacts` bounds traversal and the store's `max_bytes` bounds total selected
payload bytes. Reading rechecks the entire closure. Save the selection identifier
with the application workflow, then apply its own access and reuse decisions.
The decision functions below record host acceptance; a selection alone is not
accepted memory. The tiny dictionary above illustrates opaque preservation, not a validated
data-dict contract.

## Record acceptance and withdrawal

Applications decide whether a candidate is acceptable and who may use it. Graft
records those decisions against exact selections. A stream names one independently
reviewed subject; a new operation key represents a new review, even when its
selection is unchanged.

```{r}
path <- tempfile("artifact-decisions-")
store <- graft_artifact_store(path, create = TRUE)
report <- graft_artifact_save(
  store, "report", charToRaw("Reviewed report"), "text/plain"
)
selection <- graft_artifact_select(store, list(report))
accepted <- graft_artifact_decide(
  store, "research:topic", "review-1", expected = NULL,
  selection = selection, action = "accept", actor = "reviewer:alice",
  reason = "Supporting evidence reviewed", purpose = "research"
)
input <- graft_artifact_reuse(
  store, "research:topic", accepted$id, "research", eligible = TRUE
)
input$selection$artifacts
withdrawn <- graft_artifact_decide(
  store, "research:topic", "withdraw-1", expected = accepted$id,
  selection = selection, action = "withdraw", actor = "reviewer:alice",
  reason = "Evidence needs correction", purpose = "research"
)
graft_artifact_read_decision(store, "research:topic", accepted$id)$action
graft_artifact_read_decision(store, "research:topic")$action
unlink(path, recursive = TRUE)
```

The `eligible = TRUE` argument represents the application's current policy
check; it is not an authentication mechanism. Reuse also requires the current
accepted head, an exact purpose match and intact selected content. The earlier
acceptance remains inspectable after withdrawal, but cannot pass current reuse.
Low-level reads remain available for authorized historical inspection; applications
must enforce access on every route.

Always supply the predecessor observed during review. An identical retry returns
its original historical record and never restores eligibility. A changed retry or
stale new request fails. Withdrawal needs intact decision history but remains
possible when selected payloads are corrupt or unavailable.

Each stream uses a bounded, append-only journal. Defaults allow 1,000 decisions
and 1 MiB of aggregate journal metadata; selection bounds are independent.
Staging is not a committed decision. A complete immutable record is the commit
point, so an interrupted request can be retried without duplicating acceptance.
Concurrent writers, power-loss durability and backup recovery remain separate
operational work.

## From analysis to later reuse

Suppose Commons produces a report, table, and figure. A useful artifact layer
preserves their contents, stable identities, revisions, and references to the
exact inputs and meaning used to produce them. A correction creates a new result
while the original remains available for historical inspection.

| Concern | Proposed owner |
| --- | --- |
| Local schemas, types, constraints, and expressions | data-dict |
| Shared concepts and qualified bindings across dictionaries | A separate internal vocabulary publishing module |
| Analysis, context search, and calculation evidence | Commons |
| Artifact contents, revisions, dependencies, and saved selections | Graft shared artifact infrastructure |
| Approval, access, reuse purpose, withdrawal, and active task lifetime | The application |

The vocabulary publisher connects domain concepts across workflows. A binding
can associate two local fields with the same concept while retaining their
source dictionaries and units. It does not by itself establish that their values
can be joined or compared.

## Persistent artifacts provide material for memory

A later workflow might select an approved interpretation, a dictionary release,
and three earlier results. The saved selection records those exact revisions and
their dependencies. An unchanged day retains the full selection even when there
is no new acceptance receipt.

Saving a draft and approving its reuse are separate actions. The application
checks current eligibility when it starts a new workflow. Corrections and returning
to an earlier input revision do not automatically reapprove a withdrawn selection.
The prototype permits the application to explicitly approve it again; the
application owns who may do that and for what purpose. Historical access does not
imply permission to use that evidence in a new task.

The application also owns cancellation of active work, Reader identity, and
permanent Forget. A stored selection cannot erase copies already given to a model.

## What “replaceable” means

Applications retain exact artifacts, dependency selections, and host decisions.
They do not depend on graph tables, a schema compiler, or native store objects.
A future storage implementation must preserve those identities and history;
Graft does not yet expose a pluggable backend interface.

The earlier composition experiments compared native graph storage with immutable
manifests. Their source and observations remain in the repository as historical
research. They motivated the artifact contract and the consumer contract 3 cut;
they are not current native-store usage instructions.

Shared infrastructure remains in Graft. Tempest and Rill own product semantics,
access, approval, and retention. Publication recovery remains separate work in
[#49](https://github.com/JamesHWade/graft/issues/49).

## Use retained research in Tempest

Tempest now maps validated completed research into this artifact contract. Its
`tempest_publish_artifact_research()` retains the exact evidence closure, source
bodies, program provenance and readable report. A host explicitly accepts that
selection using `graft_artifact_decide()`.

`tempest_reuse_artifact_research()` validates the scientific evidence and consults
the current Graft decision using a host eligibility callback. New runs and
session resume recheck admission. The callback is transient; session bundles
retain content and decision provenance, never a saved permission. Historical
inspection uses `tempest_read_artifact_research()` and does not grant reuse.

The [Tempest artifact guide](https://jameshwade.github.io/tempest/articles/artifact-knowledge.html)
contains the installed offline example. Current Tempest consumer tests exercise
publication, fresh-process reads, corrections, and current admission through these
public interfaces. The earlier pinned experiment suite remains historical
evidence. These synthetic fixtures do not establish production access or erasure.
