---
title: "Persistent artifacts and selected memory"
description: >
  Save exact artifacts, record host decisions and guard current reuse.
  Understand the relationship to Commons, data-dict and the migration evidence.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent artifacts and selected memory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Your saved work should survive both the end of an R session and a change in the
software that stores it. Graft preserves reports, tables, figures and the exact
evidence selected for later workflows. Its decision journal records host reviews
and separates historical inspection from current reuse.

## Save and reopen exact content

The shared local artifact interface preserves opaque bytes and verifies them when
read. It supports trusted local files with a single writer. Application approval,
access, concurrent publication, power-loss durability and permanent erasure remain
separate requirements; saving these bytes grants no approval.

```{r}
library(graft)
path <- tempfile("artifacts-")
store <- graft_store(path, create = TRUE)
original <- graft_save(
  store, "Original report", id = "report:daily", media_type = "text/plain"
)
corrected <- graft_save(
  store, "Corrected report", id = "report:daily", media_type = "text/plain"
)
reopened <- graft_store(path)
graft_read(reopened, original)@data
graft_read(reopened, corrected)@data
unlink(path, recursive = TRUE)
```

`ArtifactRef` values contain an identity and immutable revision. Retain them with
your workflow; the store deliberately has no automatic latest or approved
selection. This project is pre-production. Consumer contract 4 uses typed S7
stores and values; the earlier graph/compiler experiments remain historical
evidence, not installed storage code.

## Pin evidence and shared meaning

A dependency is an exact artifact reference. Store dictionary and vocabulary
releases as content, then reference them alongside the evidence used in a report.
The selection preserves their bytes; it does not evaluate expressions or infer
that two measurements are comparable.

```{r}
path <- tempfile("artifact-selection-")
store <- graft_store(path, create = TRUE)
vocabulary <- graft_save(
  store, '{"concept":"sample"}',
  id = "vocabulary:lab", media_type = "application/json"
)
dictionary <- graft_save(
  store, "name: assay", id = "dictionary:assay", media_type = "application/yaml",
  dependencies = vocabulary
)
report <- graft_save(
  store, "Assay interpretation", id = "report:assay", media_type = "text/plain",
  dependencies = dictionary
)
selection <- graft_select(store, report)
reopened_selection <- graft_read_selection(graft_store(path), selection@id)
reopened_selection@artifacts
unlink(path, recursive = TRUE)
```

The returned `ArtifactSelection` includes the report, dictionary and vocabulary
revisions in its `@artifacts` list.
`max_metadata_bytes` separately bounds encoded selection metadata (1 MiB by
default). Increase it for many long artifact identities and pass that limit on
both selection creation and reading. Similarly, open the store with a larger
`max_revision_bytes` (default 1 MiB) when saving or reading artifacts whose
dependency lists produce large revision manifests.

`max_artifacts` bounds traversal and the store's `max_bytes` bounds total selected
payload bytes. Reading rechecks the entire closure. Save the selection identifier
with the application workflow, then apply its own access and reuse decisions.
The decision functions below record host acceptance; a selection alone is not
accepted memory. The tiny dictionary above illustrates opaque preservation, not a validated
data-dict contract.

## Record acceptance and withdrawal

Applications decide whether a candidate is acceptable and who may use it. Graft
records those decisions against exact selections. A stream names one independently
reviewed subject; a new operation key represents a new review, even when its
selection is unchanged.

```{r}
path <- tempfile("artifact-decisions-")
store <- graft_store(path, create = TRUE)
report <- graft_save(
  store, "Reviewed report", id = "report", media_type = "text/plain"
)
accepted <- graft_accept(
  store, report, stream = "research:topic", key = "review-1", expected = NULL,
  actor = "reviewer:alice",
  reason = "Supporting evidence reviewed", purpose = "research"
)
input <- graft_recall(
  store, "research:topic", purpose = "research", eligible = TRUE
)
input@artifacts
withdrawn <- graft_withdraw(
  store, "research:topic", expected = accepted, key = "withdraw-1",
  actor = "reviewer:alice", reason = "Evidence needs correction"
)
graft_history(store, "research:topic")[[1L]]@action
graft_history(store, "research:topic")[[2L]]@action
unlink(path, recursive = TRUE)
```

The `eligible = TRUE` argument represents the application's current policy
check; it is not an authentication mechanism. Recall also requires the current
accepted head, an exact purpose match and intact selected content. The earlier
acceptance remains inspectable through `graft_history()` after withdrawal, but
cannot pass current recall. Exact `graft_read()` calls remain available for
authorized historical inspection; applications must enforce access on every
route.

Each `Decision@selection` is an exact selection digest. `Recall@selection` is the
verified `ArtifactSelection`, while `Recall@roots` and `Recall@artifacts` contain
materialized `Artifact` values. A withdrawal derives its purpose and selection
from the verified predecessor, so callers cannot replace them with mutated
properties.

Always supply the predecessor observed during review. An identical retry returns
its original historical record and never restores eligibility. A changed retry or
stale new request fails. Withdrawal needs intact decision history but remains
possible when selected payloads are corrupt or unavailable.

Each stream uses a bounded, append-only journal. Defaults allow 1,000 decisions
and 1 MiB of aggregate journal metadata; selection bounds are independent.
Staging is not a committed decision. A complete immutable record is the commit
point, so an interrupted request can be retried without duplicating acceptance.
Concurrent writers, power-loss durability and backup recovery remain separate
operational work.

## From analysis to later reuse

Suppose Commons produces a report, table, and figure. A useful artifact layer
preserves their contents, stable identities, revisions, and references to the
exact inputs and meaning used to produce them. A correction creates a new result
while the original remains available for historical inspection.

| Concern | Proposed owner |
| --- | --- |
| Local schemas, types, constraints, and expressions | data-dict |
| Shared concepts and qualified bindings across dictionaries | A separate internal vocabulary publishing module |
| Analysis, context search, and calculation evidence | Commons |
| Artifact contents, revisions, dependencies, and saved selections | Graft shared artifact infrastructure |
| Approval, access, reuse purpose, withdrawal, and active task lifetime | The application |

The vocabulary publisher connects domain concepts across workflows. A binding
can associate two local fields with the same concept while retaining their
source dictionaries and units. It does not by itself establish that their values
can be joined or compared.

## Persistent artifacts provide material for memory

A later workflow might select an approved interpretation, a dictionary release,
and three earlier results. The saved selection records those exact revisions and
their dependencies. An unchanged day retains the full selection even when there
is no new acceptance receipt.

Saving a draft and approving its reuse are separate actions. The application
checks current eligibility when it starts a new workflow. Corrections and returning
to an earlier input revision do not automatically reapprove a withdrawn selection.
The prototype permits the application to explicitly approve it again; the
application owns who may do that and for what purpose. Historical access does not
imply permission to use that evidence in a new task.

The application also owns cancellation of active work, Reader identity, and
permanent Forget. A stored selection cannot erase copies already given to a model.

## What “replaceable” means

Applications retain exact artifacts, dependency selections, and host decisions.
They do not depend on graph tables, a schema compiler, or native store objects.
A future storage implementation must preserve those identities and history;
Graft does not yet expose a pluggable backend interface.

The earlier composition experiments compared native graph storage with immutable
manifests. Their source and observations remain in the repository as historical
research. They motivated the artifact contract and the earlier native-store cut;
they are not current native-store usage instructions.

Shared infrastructure remains in Graft. Tempest and Rill own product semantics,
access, approval, and retention. Publication recovery remains separate work in
[#49](https://github.com/JamesHWade/graft/issues/49).

## Use retained research in Tempest

Tempest now maps validated completed research into this artifact contract. Its
`tempest_publish_artifact_research()` retains the exact evidence closure, source
bodies, program provenance and readable report. A host explicitly accepts that
selection using `graft_accept()`.

`tempest_reuse_artifact_research()` validates the scientific evidence and consults
the current Graft decision using a host eligibility callback. New runs and
session resume recheck admission. The callback is transient; session bundles
retain content and decision provenance, never a saved permission. Historical
inspection uses `tempest_read_artifact_research()` and does not grant reuse.

The [Tempest artifact guide](https://jameshwade.github.io/tempest/articles/artifact-knowledge.html)
contains the installed offline example. Current Tempest consumer tests exercise
publication, fresh-process reads, corrections, and current admission through these
public interfaces. The earlier pinned experiment suite remains historical
evidence. These synthetic fixtures do not establish production access or erasure.

## PostgreSQL and host-selected scopes

`graft_store_postgres()` uses the same artifact, selection, decision,
and vocabulary interfaces inside a host-owned DBI transaction. Its optional
DBI and RPostgres dependencies are needed only for PostgreSQL stores.

```{r, eval=FALSE}
DBI::dbWithTransaction(connection, {
  # The host derives this scope from authenticated identity and checks access.
  store <- graft_store_postgres(
    connection, scope = trusted_scope, create = TRUE
  )
  ref <- graft_save(store, "Accepted text", id = "note", media_type = "text/plain")
  accepted <- graft_accept(
    store, ref, stream = "note", key = "review-1", expected = NULL,
    actor = trusted_actor,
    reason = "Explicit review", purpose = "reader-context"
  )
  # Host catalog and receipt writes can use this same connection.
})
# Only successful transaction completion commits the operation.
```

Every retained row belongs to one scope. Graft serializes operations within a
scope until commit or rollback; the host can atomically retain its own records
in the same transaction. A scope key is not authentication, and the database
role can still access other scopes directly. Applications must check access
before every operation and supply scope from trusted identity, including in
background workers. Connections and store handles are process-local.

This boundary does not implement permanent Forget or prevent an old database
backup from restoring deleted content. Those remain explicit host and storage
obligations before a broad memory rollout.

## Build a verified replacement

A host can preview exclusions with `graft_plan_replacement()` and copy exact
survivors into a separate empty store with `graft_replace()`.
`graft_manifest()` verifies the complete logical object set, including
historical decisions. See [Verify a replacement artifact store](artifact-recovery.html)
for the runnable example and the separate Forget and restore obligations.

For a complete copy that retains every object and historical decision, use
`graft_backup()`. Verification and restore require the receipt retained
by the host independently of the bundle. See [Back up and restore an artifact store](artifact-backups.html).
