---
title: "Persistent artifacts and selected memory"
description: >
  Save exact artifacts, record host decisions and guard current reuse.
  Understand the relationship to Commons, data-dict and the migration evidence.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent artifacts and selected memory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Saved work should remain available after an R session ends and after the
software that stores it changes. Graft keeps reports, tables, figures, and the
exact evidence selected for later workflows. Its decision journal records host
reviews and distinguishes historical inspection from current reuse.

## Save and reopen exact content

The shared local artifact interface preserves opaque bytes and checks them when
they are read. It supports trusted local files with one writer. Applications
still need their own rules for approval, access, concurrent publication,
power-loss durability, and permanent erasure. Saving bytes does not approve
them for use.

```{r}
library(graft)
path <- tempfile("artifacts-")
store <- graft_store(path, create = TRUE)
original <- graft_save(
  store, "Original report", id = "report:daily", media_type = "text/plain"
)
corrected <- graft_save(
  store, "Corrected report", id = "report:daily", media_type = "text/plain"
)
reopened <- graft_store(path)
graft_read(reopened, original)@data
graft_read(reopened, corrected)@data
unlink(path, recursive = TRUE)
```

`ArtifactRef` values contain an identity and immutable revision. Keep them with
the workflow that uses them. The store has no automatic latest or approved
selection. This project is pre-production. Consumer contract 4 uses typed S7
stores and values. The earlier graph/compiler experiments remain historical
evidence, not installed storage code.

## Pin evidence and shared meaning

A dependency is an exact artifact reference. Save dictionary and vocabulary
releases as content, then reference them alongside the evidence for a report.
The selection preserves their bytes. It does not evaluate expressions or infer
that two measurements are comparable.

```{r}
path <- tempfile("artifact-selection-")
store <- graft_store(path, create = TRUE)
vocabulary <- graft_save(
  store, '{"concept":"sample"}',
  id = "vocabulary:lab", media_type = "application/json"
)
dictionary <- graft_save(
  store, "name: assay", id = "dictionary:assay", media_type = "application/yaml",
  dependencies = vocabulary
)
report <- graft_save(
  store, "Assay interpretation", id = "report:assay", media_type = "text/plain",
  dependencies = dictionary
)
selection <- graft_select(store, report)
reopened_selection <- graft_read_selection(graft_store(path), selection@id)
reopened_selection@artifacts
unlink(path, recursive = TRUE)
```

The returned `ArtifactSelection` includes the report, dictionary, and vocabulary
revisions in `@artifacts`.
`max_metadata_bytes` separately limits encoded selection metadata (1 MiB by
default). Increase it when artifact identities are numerous or long, and pass
the same limit when creating and reading the selection. Likewise, open the store
with a larger `max_revision_bytes` (default 1 MiB) when dependency lists create
large revision manifests.

`max_artifacts` limits traversal, and the store's `max_bytes` limits the total
selected payload bytes. Reading checks the full dependency chain again. Save the
selection identifier with the application workflow, then apply the application's
access and reuse decisions. The decision functions below record application
approval; a selection alone is not accepted memory. The small dictionary above
shows that Graft keeps supplied bytes without interpreting them. It does not
demonstrate a validated data-dict contract.

## Record acceptance and withdrawal

Applications decide whether a candidate is acceptable and who may use it. Graft
records each decision against an exact selection. A stream names one independently
reviewed subject. A new operation key starts a new review even when the selection
has not changed.

```{r}
path <- tempfile("artifact-decisions-")
store <- graft_store(path, create = TRUE)
report <- graft_save(
  store, "Reviewed report", id = "report", media_type = "text/plain"
)
accepted <- graft_accept(
  store, report, stream = "research:topic", key = "review-1", expected = NULL,
  actor = "reviewer:alice",
  reason = "Supporting evidence reviewed", purpose = "research"
)
input <- graft_recall(
  store, "research:topic", purpose = "research", eligible = TRUE
)
input@artifacts
withdrawn <- graft_withdraw(
  store, "research:topic", expected = accepted, key = "withdraw-1",
  actor = "reviewer:alice", reason = "Evidence needs correction"
)
graft_history(store, "research:topic")[[1L]]@action
graft_history(store, "research:topic")[[2L]]@action
unlink(path, recursive = TRUE)
```

The `eligible = TRUE` argument stands in for the application's current check that
the material may be used; it does not authenticate anyone. Recall also requires
the current accepted head, an exact purpose match, and intact selected content.
The earlier acceptance remains available through `graft_history()` after
withdrawal, but it cannot pass current recall. Authorized historical inspection
can still use exact `graft_read()` calls. Applications must enforce access on
every route.

Each `Decision@selection` is an exact selection digest. `Recall@selection` is the
verified `ArtifactSelection`, while `Recall@roots` and `Recall@artifacts` contain
materialized `Artifact` values. A withdrawal derives its purpose and selection
from the verified predecessor, so callers cannot replace them with changed
properties.

Always supply the predecessor observed during review. An identical retry returns
its original historical record and does not restore eligibility. A changed retry
or a stale new request fails. Withdrawal needs intact decision history, but it can
still proceed when selected payloads are corrupt or unavailable.

Each stream uses a bounded, append-only journal. The defaults allow 1,000
decisions and 1 MiB of aggregate journal metadata; selection bounds are
independent. Staging is not a committed decision. A complete immutable record is
the commit point, so an interrupted request can be retried without duplicating
acceptance. Concurrent writers, power-loss durability, and backup recovery are
separate operational work.

## From analysis to later reuse

Suppose Commons produces a report, table, and figure. The artifact layer can
keep their contents, stable identities, revisions, and references to the exact
inputs and meaning used to produce them. A correction creates a new result while
the original remains available for historical inspection.

| Concern | Proposed owner |
| --- | --- |
| Local schemas, types, constraints, and expressions | data-dict |
| Shared concepts and qualified bindings across dictionaries | A separate internal vocabulary publishing module |
| Analysis, context search, and calculation evidence | Commons |
| Artifact contents, revisions, dependencies, and saved selections | Graft shared artifact infrastructure |
| Approval, access, reuse purpose, withdrawal, and active task lifetime | The application |

The vocabulary publisher connects domain concepts across workflows. A binding
can associate two local fields with one concept while retaining their source
dictionaries and units. It does not establish that their values can be joined or
compared.

## Persistent artifacts provide material for memory

A later workflow might select an approved interpretation, a dictionary release,
and three earlier results. The saved selection records those exact revisions and
dependencies. The full selection remains available even when no new acceptance
receipt is recorded.

Saving a draft and approving reuse are separate actions. When a new workflow
starts, the application checks current eligibility. A correction or a return to
an earlier input revision does not reapprove a withdrawn selection. The prototype
lets the application approve it again explicitly. The application decides who
may do that and for what purpose. Historical access does not grant permission to
use the evidence in a new task.

The application also owns cancellation of active work, Reader identity, and
permanent Forget. A stored selection cannot erase copies already given to a
model.

## What replaceable means

Applications retain exact artifacts, dependency selections, and application decisions.
They do not depend on graph tables, a schema compiler, or native store objects.
A future storage implementation must keep those identities and history. Graft
does not yet expose a pluggable backend interface.

Earlier composition experiments compared native graph storage with immutable
manifests. Their source and observations remain in the repository as historical
research. They led to the artifact contract and the retirement of the native
store. Their examples document that earlier architecture and are not instructions
for the current store.

Graft keeps the shared infrastructure. Tempest and Rill own product semantics,
access, approval, and retention. Publication recovery remains separate work in
[#49](https://github.com/JamesHWade/graft/issues/49).

## Use retained research in Tempest

Tempest maps validated completed research into this artifact contract. Its
`tempest_publish_artifact_research()` retains the exact evidence and all of its
dependencies, along with source bodies, program provenance, and a readable
report. The application explicitly accepts that selection with `graft_accept()`.

`tempest_reuse_artifact_research()` validates the scientific evidence and checks
the current Graft decision through an application-provided eligibility callback.
New runs and resumed sessions check permission to reuse the evidence again.
The callback is transient. Session bundles retain content and decision
provenance, never a saved permission. Historical
inspection uses `tempest_read_artifact_research()` and does not grant reuse.

The [Tempest artifact guide](https://jameshwade.github.io/tempest/articles/artifact-knowledge.html)
contains the installed offline example. Current Tempest consumer tests exercise
publication, fresh-process reads, corrections, and current admission through
these public interfaces. The earlier pinned experiment suite remains historical
evidence. These synthetic fixtures do not establish production access or
erasure.

## PostgreSQL and application scopes

`graft_store_postgres()` provides the same artifact, selection, decision, and
vocabulary interfaces inside a DBI transaction controlled by the application. Its
optional DBI and RPostgres dependencies are needed only for PostgreSQL stores.

```{r, eval=FALSE}
DBI::dbWithTransaction(connection, {
  # The host derives this scope from authenticated identity and checks access.
  store <- graft_store_postgres(
    connection, scope = trusted_scope, create = TRUE
  )
  ref <- graft_save(store, "Accepted text", id = "note", media_type = "text/plain")
  accepted <- graft_accept(
    store, ref, stream = "note", key = "review-1", expected = NULL,
    actor = trusted_actor,
    reason = "Explicit review", purpose = "reader-context"
  )
  # Host catalog and receipt writes can use this same connection.
})
# Only successful transaction completion commits the operation.
```

Every retained row belongs to one scope. Graft serializes operations within a
scope until commit or rollback, and the application can retain its own records
atomically in the same transaction. A scope key does not authenticate anyone;
the database role can still access other scopes directly. Applications must check
access before every operation and derive the scope from trusted identity,
including in background workers. Connections and store handles are process-local.

This interface does not implement permanent Forget or stop an old database backup
from restoring deleted content. The application and storage system must handle those
obligations before a broad memory rollout.

## Build a verified replacement

An application can preview exclusions with `graft_plan_replacement()` and copy
the remaining artifacts into a separate empty store with `graft_replace()`,
preserving their exact contents.
`graft_manifest()` verifies the complete logical object set, including historical
decisions. See [Verify a replacement artifact store](artifact-recovery.html)
for the runnable example and the separate Forget and restore obligations.

For a complete copy that retains every object and historical decision, use
`graft_backup()`. Verification and restore require the receipt that the
application keeps separately from the bundle. See [Back up and restore an artifact store](artifact-backups.html).
