---
title: "Keep Readers' accepted knowledge separate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Keep Readers' accepted knowledge separate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(graft)
```

Two Readers may accept different interpretations of the same article. A later
task must use the authenticated Reader's accepted knowledge, even if another
Reader has the same record ID. This offline example opens a separate Graft
store for each Reader and checks host authorization before every read.

This is an application-owned teaching recipe for
[#47](https://github.com/JamesHWade/graft/issues/47). It uses synthetic records
and grants, not Rill authentication. The
[storage and access proposal](https://github.com/JamesHWade/graft/blob/main/adr/0005-isolate-reader-knowledge-stores.md)
compares separate stores with shared scoped reads and specifies the remaining
Rill work. No Graft access-control API is added.

## Bind access in the host

The host supplies the Reader, store locator, and current grant revision. These
inputs must never come from model arguments or a saved reuse basis. The recipe
opens read-only, checks the expected store UUID, collects the read, rechecks the
grant revision, and closes the connection. It returns a generic error when any
step fails.

```{r bind}
example <- new.env()
sys.source(system.file("examples/reader-access.R", package = "graft"), example)
directory <- tempfile("reader-access-")
dir.create(directory)
locations <- example$reader_access_fixture(directory)
grants <- new.env()
grants$alice <- "alice-grant-1"
grants$bob <- "bob-grant-1"
bind <- function(reader) {
  example$reader_knowledge_access(
    reader,
    locate = \(reader) locations[[reader]],
    grant = \(reader) grants[[reader]]
  )
}
alice <- bind("alice")
bob <- bind("bob")
alice(\(store) graft_get(store, "knowledge:preference"))$record$body
bob(\(store) graft_get(store, "knowledge:preference"))$record$body
```

The colliding ID resolves inside the selected store. A foreign ID and an
unavailable ID produce the same boundary error. Search and accepted
calculations only observe the selected Reader's data.

```{r scope}
unavailable <- function(read) {
  tryCatch(read(), reader_knowledge_unavailable = conditionMessage)
}
unavailable(\() alice(\(store) graft_get(store, "knowledge:interpretation:bob")))
unavailable(\() alice(\(store) graft_get(store, "absent")))
alice(\(store) nrow(graft_find(store, "bob")))
alice(\(store) graft_calculate(store, "knowledge_count"))
bob(\(store) graft_calculate(store, "knowledge_count"))
```

Separate files matter here: a column marked `display: restricted` only omits
that column. It does not authorize a Reader, filter other records, or protect
aggregates in a shared store. The host's callback is trusted R code; do not
expose it, filesystem access, or an R evaluator to the model.

## Pin a tool without freezing permission

The tool exposes only a record ID. It obtains its complete JSON result and
receipt from `graft_tools()` after reopening the authorized store at the pinned
boundary. Revoking access after construction prevents the next call.

```{r tool}
tool <- example$reader_record_tool(alice, locations$alice$snapshot)
value <- jsonlite::fromJSON(tool("knowledge:preference"))
value$result$record$body
identical(value$receipt$store$id, locations$alice$store_id)
grants$alice <- NULL
unavailable(\() tool("knowledge:preference"))
bob(\(store) graft_get(store, "knowledge:preference"))$record$body
```

The same callback rejects a changed grant revision during a read, including a
revoke/regrant cycle. Content already delivered to a model or client remains a
host lifecycle concern: cancel affected runs and discard their reusable context.
Permanent erasure and backup restore remain separate work in
[#48](https://github.com/JamesHWade/graft/issues/48).

## Reconnect with a complete basis

Persist the [complete selected basis](reuse-basis.html), then reconstruct the
Reader's access from trusted host state in the new session or worker. The
checkpoint contains no locator or credentials. A basis from another store fails
even when its selected record ID exists locally.

```{r rebind}
grants$alice <- "alice-grant-2"
sys.source(system.file("examples/reuse-basis.R", package = "graft"), example)
basis <- alice(\(store) example$capture_reuse_basis(
  store,
  "knowledge:preference",
  data.frame(outcome = character(), dependency = character())
))
checkpoint <- file.path(directory, "basis.rds")
saveRDS(basis, checkpoint)
restored <- readRDS(checkpoint)
reconnected <- bind("alice")
reconnected(\(store) example$read_reuse_basis(store, restored, \(ids) TRUE))
unavailable(\() bob(\(store) example$read_reuse_basis(store, restored, \(ids) TRUE)))
```

Here the inner callback permits the complete selection because the outer access
boundary already authorizes this Reader's store. A product must additionally
apply its current consultation-eligibility policy. Package tests repeat the
rebind in a separate R process, with all writer connections closed first.

The example proves isolated Graft reads and host tool composition. It does not
establish durable hosting, concurrent writer access, Rill's Document hydration,
or permanent Forget. Those gates remain explicit in
[#51](https://github.com/JamesHWade/graft/issues/51) and the access proposal.

```{r cleanup, include=FALSE}
unlink(directory, recursive = TRUE)
```
