---
title: "Retrieve accepted knowledge"
description: >
  Get current records, search public fields, inspect accepted history, and
  expose the same bounded reads to agent hosts.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve accepted knowledge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Graft provides four read functions. Current reads use the active contract;
`graft_history()` uses the exact contract recorded for each revision. None
exposes raw SQL or a mutation path.

| Need | Function | Result |
|---|---|---|
| One current record | `graft_get()` | The record and its accepted context |
| Search public fields | `graft_find()` | Ranked, bounded matches |
| A fixed advanced operation | `graft_query()` | A validated operation-specific result |
| Accepted revisions | `graft_history()` | Newest-first immutable history |
| Read-only agent access | `graft_tools()` | Tool definitions backed by the same reads |

## Create some accepted knowledge

This example uses the resolved data-dict contract included with Graft. It
describes people, organizations, and employment records that refer to both.
The foreign keys are checked during planning, but they are not semantic graph
edges.

```{r setup}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "team-directory.data-dict.json",
  package = "graft",
  mustWork = TRUE
))
store <- graft_open(schema, ":memory:", okf = "disabled")

graft_ingest(
  store,
  list(
    organization = data.frame(
      id = "org:daily-planet",
      name = "Daily Planet"
    ),
    person = data.frame(
      id = "person:lois-lane",
      full_name = "Lois Lane",
      job_title = "Reporter"
    ),
    employment = data.frame(
      id = "employment:lois-lane:daily-planet",
      person_id = "person:lois-lane",
      organization_id = "org:daily-planet"
    )
  ),
  graft_provenance(
    producer = "team-directory-import",
    idempotency_key = "team-directory-v1"
  )
)
```

## Get one current record

Use `graft_get()` when the stable identifier is already known:

```{r get}
person <- graft_get(store, "person:lois-lane")
person$record
```

The result includes its class and accepted context. A missing identifier
raises a typed package error; callers do not need to know how the record is
stored internally.

## Find records by public text

Use `graft_find()` when you know a name or phrase rather than an identifier:

```{r find}
graft_find(store, "Lois", class = "person", limit = 10)
graft_find(store, "Daily", class = "organization", limit = 10)
```

The compiled contract determines which fields are searchable and which are
sensitive. An optional class restriction narrows the search; every call remains
bounded by its result limit.

## Recover accepted history

Accept a later version of the same person:

```{r update}
graft_ingest(
  store,
  list(
    person = data.frame(
      id = "person:lois-lane",
      full_name = "Lois Lane",
      job_title = "Investigative reporter"
    )
  ),
  graft_provenance(
    producer = "team-directory-import",
    idempotency_key = "team-directory-v2"
  )
)
```

`graft_history()` returns the accepted revisions in newest-first order:

```{r history}
history <- graft_history(
  store,
  id = "person:lois-lane",
  limit = 100
)

history[, c("batch_id", "changed_fields", "record")]
```

An accepted batch ID or `POSIXt` value selects state at an exact commit
boundary. Commit order, rather than a timestamp inside the domain record,
defines that boundary.

## Use fixed advanced operations

`graft_query()` accepts named operations with validated request shapes. For
example, the integrity operation checks the revision chain and, when
requested, its derived projections:

```{r integrity}
graft_query(
  store,
  operation = "integrity",
  request = list(projections = TRUE),
  limit = 100
)
```

The `neighbors` operation returns semantic edges only when the active contract
declares graph-producing semantic statements or edges. Graft does not turn the
team directory's data-dict foreign keys, or an ordinary LinkML object-reference
slot, into traversal edges. See [Add graph semantics with
LinkML](linkml-schema.html) for that contract shape.

Other fixed operations cover exact identifiers, claims, evidence, unresolved
mentions, and bounded neighbors where the contract supports them. Unknown
request members, unbounded traversal, and arbitrary SQL are rejected.

## Give agents the same bounded reads

```{r tools}
tools <- graft_tools(store)
names(tools)
```

The definitions call `graft_find()`, `graft_get()`, `graft_query()`, and
`graft_history()`. They expose no write operation, raw database connection,
filesystem access, or network access. The host decides which provider receives
them and remains responsible for tool authorization.

Every operation reports its applicable limits and truncation state so a host
can distinguish a complete result from a bounded prefix.

```{r close}
graft_close(store)
```

Read [Change control](knowledge-change-control.html) for the path from proposal
to accepted revision and [Architecture](architecture.html) for how the ledger
and derived views fit together.
