---
title: "Retrieve accepted knowledge"
description: >
  Choose the smallest bounded read operation for current records, search,
  advanced projections, immutable history, or agent tools.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve accepted knowledge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

graft's read surface is deliberately smaller than its storage model. Four
functions cover current records, search, fixed advanced operations, and
immutable history. `graft_tools()` exposes the same accepted reads to an agent
host without adding mutation or raw SQL.

| Need | Function | Result |
|---|---|---|
| One current record | `graft_get()` | A hydrated record and its accepted context |
| Search public fields | `graft_find()` | Ranked, bounded matches |
| A fixed advanced shape | `graft_query()` | A validated operation-specific result |
| Accepted revisions | `graft_history()` | Newest-first immutable history |
| Read-only agent access | `graft_tools()` | Tool definitions backed by the four reads |

All reads are governed by the active contract and accepted revision ledger.
Internal projections make them efficient, but projections are not independent
sources of knowledge.

## Create a small accepted graph

```{r setup}
library(graft)

schema <- graft_schema(system.file(
  "extdata",
  "personinfo.graft.json",
  package = "graft"
))
store <- graft_open(schema, ":memory:", okf = "disabled")

graft_ingest(
  store,
  list(
    Organization = data.frame(
      id = "org:daily-planet",
      name = "Daily Planet"
    ),
    Person = data.frame(
      id = "person:clark-kent",
      full_name = "Clark Kent",
      aliases = I(list(c("Superman", "Kal-El"))),
      employed_by = I(list("org:daily-planet"))
    )
  ),
  graft_provenance(
    producer = "directory-import",
    idempotency_key = "daily-planet-v1"
  )
)
```

## Get one current record

Use `graft_get()` when the stable identifier is already known:

```{r get}
person <- graft_get(store, "person:clark-kent")
person$record
```

The result includes accepted context rather than returning an unqualified
storage row. Missing identifiers return a typed package error; callers do not
need to know which projection table serves the request.

## Find records by public text

Use `graft_find()` for human-facing lookup across public search fields declared
by the LinkML contract:

```{r find}
graft_find(store, "Clark", class = "Person", limit = 10)
graft_find(store, "Daily", class = "Organization", limit = 10)
```

Class and limit are explicit. Sensitive fields are excluded by contract rather
than filtered by the caller after retrieval.

## Query a fixed advanced operation

`graft_query()` is the extension point for reads that need a richer shape. The
operation chooses a validated request schema; the request supplies only values
for that operation.

```{r neighbors}
graft_query(
  store,
  operation = "neighbors",
  request = list(
    id = "person:clark-kent",
    projection = "semantic",
    hops = 1,
    max_nodes = 25,
    max_edges = 50
  )
)
```

Operations cover exact identifiers, identifiers, claims, evidence, bounded
neighbors, unresolved mentions, and integrity. Unknown request members,
unbounded traversal, and arbitrary SQL are rejected.

Use the integrity operation when diagnosing the accepted ledger or derived
views:

```{r integrity}
graft_query(
  store,
  operation = "integrity",
  request = list(projections = TRUE),
  limit = 100
)
```

Projection drift is repairable from accepted revisions. A successful current
query is not treated as proof that the historical chain is intact; integrity
inspection checks the authoritative chain directly.

## Recover immutable history

Use `graft_history()` when the question includes "when," "why," or "what
changed":

```{r history}
history <- graft_history(
  store,
  id = "person:clark-kent",
  limit = 100
)

history[, c("batch_id", "changed_fields", "record")]
```

An accepted batch ID or `POSIXt` value selects state at an exact commit
boundary. Commit order, not a domain record's timestamp, defines historical
boundaries.

## Give agents the same bounded reads

```{r tools}
tools <- graft_tools(store)
names(tools)
```

The returned definitions are backed by `graft_find()`, `graft_get()`,
`graft_query()`, and `graft_history()`. They expose no write operation, raw
database connection, filesystem access, or network access. The host decides
which provider receives them and remains responsible for tool authorization.

Every operation reports its applicable limits and truncation state so a host
can distinguish "complete result" from "bounded prefix."

```{r close}
graft_close(store)
```

Read [Architecture](architecture.html) for the ledger/projection boundary and
[Change control](knowledge-change-control.html) for how a record becomes part
of the accepted view.
