---
title: "Retain an exact reuse basis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retain an exact reuse basis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(graft)
```

A snapshot records an accepted boundary. A reuse basis additionally records the
exact records a task selected and the evidence it needs. Keep that complete
selection on an unchanged day; an empty change receipt is not an empty basis.

This example uses the [narrative teaching data](ecosystem.html). The functions
sourced below are application-owned recipes, not exported Graft interfaces or a
supported checkpoint format. Tempest and Rill retain their own selection,
authorization, evidence, and lifecycle semantics.

## Capture the complete selection

```{r capture}
example <- new.env()
sys.source(system.file(
  "examples/narrative-knowledge.R", package = "graft"
), example)
sys.source(system.file(
  "examples/reuse-basis.R", package = "graft"
), example)
store <- graft_open(graft_schema(system.file(
  "extdata/narrative-knowledge.data-dict.json", package = "graft"
)), okf = "disabled")
accepted <- graft_ingest(store, example$narrative_records(),
  graft_provenance("teaching-host", idempotency_key = "seed"))

dependencies <- data.frame(
  outcome = c("knowledge:interpretation", "support:interpretation"),
  dependency = c("support:interpretation", "source:trial-v1")
)
basis <- example$capture_reuse_basis(
  store,
  roots = c("knowledge:interpretation", "knowledge:preference"),
  dependencies = dependencies
)
basis$records[, c("record_id", "class")]
```

The basis contains both root outcomes and their required support/source records.
The preference has no fabricated citation. The application declares this closure;
it must validate its own source versions and anchors. The recipe caps the whole
selection at 1,000 records, preserving Tempest's current ceiling. It stops if the
closure is too large instead of quietly dropping evidence.

The checkpoint stores a snapshot and exact revision references, with no connection
or access grant. Restore trusted host checkpoints in a fresh process by opening
the intended store, establishing current permission, and then calling
`read_reuse_basis()`. Never import untrusted RDS or persist policy functions in a
model-authored checkpoint.

```{r exact}
# Synthetic local host authorization. Real applications decide from current identity.
allow_read <- \(ids) all(ids %in% basis$records$record_id)
before <- example$read_reuse_basis(store, basis, allow_read)
length(before)
example$review_reuse_basis(store, basis)$needs_review
```

## Review changes without losing the old meaning

```{r changed}
source <- example$narrative_records()$source
source$quote <- "The larger synthetic trial supports only a preliminary conclusion."
correction <- graft_ingest(store, list(source = source),
  graft_provenance("teaching-host", idempotency_key = "source-correction"))
assessment <- example$review_reuse_basis(store, basis)
assessment$changes[, c("record_id", "action")]
assessment$needs_review
identical(example$read_reuse_basis(store, basis, allow_read), before)
```

The interpretation needs review because its declared evidence changed. The
preference does not. The old interpretation and exact evidence remain available
for authorized inspection; the review signal neither proves the interpretation
false nor rewrites its content.

`graft_changes()` now filters IDs **before** its result limit:

```{r scoped-changes}
changes <- graft_changes(
  store, since = basis$snapshot,
  record_ids = basis$records$record_id, limit = 1000L
)
stopifnot(!isTRUE(attr(changes, "truncated")))
```

Unrelated changes cannot crowd this selection out. A selected deletion still
appears by ID even when its returned record is `NULL`. `NULL` IDs select the
whole store; `character()` selects none. Unknown IDs return no changes, so this
query is not a completeness check. Read the entire exact basis separately and
fail if any selected revision is unavailable. Discovery of new relevant records
is a separate operation whose completeness the host must check.

## Enforce eligibility at the host boundary

Acceptance, historical access, current eligibility, task relevance and factual
truth are separate decisions. The teaching host exposes one no-argument tool
that can only consult its selected bundle. Current policy is checked before and
after reading; a denied dependency prevents the entire result from being emitted.

```{r consultation, eval=FALSE}
run_epoch <- host$current_policy_epoch()
tool <- example$reuse_consultation_tool(
  store, basis,
  allow_read = \(ids) host$can_read_all(ids),
  eligible = function(basis) {
    identical(run_epoch, host$current_policy_epoch()) &&
      host$may_consult(basis$roots)
  }
)
chat$set_tools(list(consult_selected_knowledge = tool))
```

Here `host` is illustrative application code, not a Graft object. The host owns
Reader identity, Archive/restore, supersession, review decisions and policy
invalidation. A revision can remain historically readable while being ineligible
for automatic consultation. A direct request or established host policy can
also authorize acceptance without a new click for every operation.

Expose no unrestricted store tools, history, SQL, Commons copy, or general R
executor alongside this restricted tool. If an application exposes any of those
paths, it must enforce the same policy there. This custom tool returns selected
data; it does not manufacture a canonical Graft receipt or verification label.

Archive, permission changes, supersession and relevant dependency changes must
also invalidate affected active runs and stale context. Restore permits a fresh
authorized run. A refusal on the next tool call cannot remove data already sent
to a model; cancel or rebuild the affected conversation before continuing.

Permanent Forget is stronger: old snapshots, checkpoints, caches and backups
must not restore erased content. The teaching callback proves host refusal, not
physical erasure. Graft's permanent purge and restore contract remains
[#48](https://github.com/JamesHWade/graft/issues/48), and Reader access remains
[#47](https://github.com/JamesHWade/graft/issues/47).

## Consumer status

The offline tests cover research and reading selections with real Graft stores,
a separate worker, and a real ellmer loop served by local fixtures. They do not
claim completed Tempest or Rill integration. Tempest's current public schema
loader requires Graft contract 0.2.x; this slice reports 0.5.0. The contract update
and actual product proof belong to [#50](https://github.com/JamesHWade/graft/issues/50).
Rill's runtime proof remains [#51](https://github.com/JamesHWade/graft/issues/51).

The detailed revision, supersession and exact-reference contract is recorded in
[#45](https://github.com/JamesHWade/graft/issues/45) and
[#46](https://github.com/JamesHWade/graft/issues/46).

```{r close, include=FALSE}
graft_close(store)
```
