---
title: "Shared concepts across workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shared concepts across workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Graft publishes a shared vocabulary alongside persistent artifacts. Data-dict
owns each workflow's table contract; Commons can use the published context and
local dictionaries when running a workflow. Plain R can consume the same release.

## Publish one release

The bundled example describes two laboratories that use different names, units,
and measurement conditions. Concepts identify what their fields describe.
Relationships state a direction with explicit domain and range concepts. Qualified
bindings preserve each workflow's table, field, identity scope, grain, and condition.

```{r}
library(graft)
path <- system.file("examples", "vocabulary", "v1", "bindings.json", package = "graft")
store <- graft_artifact_store("lab-vocabulary", create = TRUE)
selection <- graft_vocabulary_publish(store, path)
release <- graft_vocabulary_read(store, selection)
cat(paste(release$context, collapse = "\n"))
```

Publishing requires `datadict` and its data-dict binary. Graft calls the public
`validate-spec` and `export-spec` commands against captured dictionary bytes. It
preserves the exact companion JSON, vocabulary JSON, dictionary YAML, resolved
exports, validation provenance (including the binary digest), and generated
Markdown in one artifact selection. Each source file is limited to 1 MiB and must
be self-contained. A release can contain at most 498 dictionaries; its sources and
generated outputs must fit the artifact store's aggregate byte limit. Graft checks
the dictionary count and source sizes before running data-dict. It does not parse
dictionary YAML or evaluate expressions.

Reads verify the entire artifact closure and rebuild the context from the retained
release. They need neither the original files nor an installed data-dict binary.
This is content integrity, not a signature or independent proof of the publisher's
identity or claims. Stores retain the trusted local, single-writer limits described
in [persistent artifacts](persistent-artifacts.html).

## Companion format

A `graft-bindings/1` JSON file has six fields:

| Field | Meaning |
|---|---|
| `format` | `graft-bindings/1` |
| `release` | Caller-owned release identifier |
| `vocabulary` | Vocabulary `release`, sibling `path`, and SHA-256 `sha256` |
| `dictionaries` | Array of dictionary release `id`, `workflow`, sibling `path`, and `sha256` references |
| `bindings` | Array of concept or relationship bindings |
| `assertions` | Array of explicitly qualified assertions, which may be empty |

Vocabulary files use `graft-vocabulary/1`, a `release`, and an array of `terms`.
Every term has a stable `id`, `kind` (`concept` or `relationship`), `definition`, and
an array of `aliases`. Relationships additionally name `domain` and `range` concept
IDs. Aliases are descriptive labels; they do not create equivalence or resolution rules.

Every binding has `id`, `kind`, `dictionary`, `table`, `term`, `scope`, `grain`, and
`condition`. A concept binding names one `field`; a relationship binding names
`from` and `to` fields. Each endpoint must have a concept binding matching the
relationship's domain or range and its scope, grain, and condition. Multiple
concepts for one dictionary/table/field are rejected, even if differently qualified.

Every assertion has `id`, `subject`, `predicate`, `object`, `source`, `status`,
`negated`, `time`, and `scope`. The predicate must name a declared relationship.
Status is `draft`, `accepted`, or `retracted`; negation is Boolean. Subjects and
objects are opaque application identities: their actual membership in domain and
range is not checked. These values are publisher statements, not Graft decisions.
Time and condition are descriptive strings, not executable rules.

See the [complete example files](https://github.com/JamesHWade/graft/tree/main/inst/examples/vocabulary)
for every field. Missing, unknown, duplicate, or ambiguous bindings fail publication.

## Use the same release in R and Commons

Plain R can inspect a binding and select its local field without guessing a join:

```{r}
bindings <- Filter(function(x) x$kind == "concept" &&
  x$term == "urn:example:temperature", release$bindings$bindings)
lapply(bindings, function(x) x[c("dictionary", "table", "field", "scope", "grain", "condition")])
```

A Commons caller supplies the same generated context alongside its own dictionary:

```{r}
context_file <- tempfile(fileext = ".md")
writeLines(release$context, context_file)
context <- commons::context_layer(files = context_file)
source <- commons::data_source(
  readings = data.frame(reading_id = "r1", sample_id = "s1", temperature_c = 20),
  dictionary = file.path(dirname(path), "lab-a.yaml")
)
# Supply context and source to the Commons host; retain release$references
# and selection with resulting artifacts.
```

The integration test constructs a real Commons object with the published context;
this checks constructor consumption, not model retrieval quality or factual reasoning.
Shared concepts do not establish safe joins, comparable units, access permission,
or execution authority. Negated and retracted assertions remain qualified data.
There is no ontology reasoner, unit converter, new expression compiler, or automatic
procedure selection.

## Retain earlier releases

Publish changed dictionaries and bindings as a new release and keep both selection
digests. A historical reader resolves exact bytes, never a mutable latest pointer.
Publication itself does not accept the release: hosts can use Graft's decision
journal when review and current eligibility are needed.

```{r}
next_path <- system.file("examples", "vocabulary", "v2", "bindings.json", package = "graft")
next_selection <- graft_vocabulary_publish(store, next_path)
original <- graft_vocabulary_read(store, selection)
updated <- graft_vocabulary_read(store, next_selection)
```
